---
title: FAS6337C - Final Exam
author: Marcel Gietzmann-Sanders
---

```{r}
library(lme4)

col2rgbA<-function(color,transparency)
{
  rgb(t(col2rgb(color))/255,alpha=transparency)
}
```

```{r}
setwd("/workspaces/schooling/population_dynamics/final/") 
data <- read.table("data/King_Mackerel_sim-1.csv", header=T, sep=",") 
data$Age <- as.numeric(data$Age)
head(data)
```

# 1. Fit a length-weight relationship to the pooled (regardless of sex) data.

```{r}
length_weight <- na.omit(data[,c('TL', 'Wt')])
length_weight$lTL <- log(length_weight$TL)
length_weight$lWt <- log(length_weight$Wt)
```

```{r}
plot(
    length_weight$lTL, 
    length_weight$lWt, 
    xlab="log(Length)", 
    ylab="log(Weight)", 
    main="Length-Weight Relationship"
)
```

```{r}
(fit <- lm(lWt ~ lTL, data=length_weight))
```

## a. Report the length-weight parameters

```{r}
a <- exp(coef(fit)[1])
b <- coef(fit)[2]
c(a, b)
```

## b. Plot the mean weight at age against the data
```{r}
length_weight$predWt <- a * length_weight$TL^b
plot(
    length_weight$TL, 
    length_weight$Wt, 
    xlab="Length (mm)", 
    ylab="Weight (kg)", 
    main="Length-Weight Relationship"
)
lines(length_weight$TL[order(length_weight$TL)], length_weight$predWt[order(length_weight$TL)], col="red")
```

# 2. Fit a series of von Bertalanffy growth models to the King Mackerel data. Determine if males and females exihibit secually dimorphic growth.

```{r}
filtered_data <- na.omit(data[, c('Age', 'TL', 'Sex')])
head(filtered_data)
```

```{r}
plot(
    filtered_data$Age, 
    filtered_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship"
)
```

```{r}
predict_length <- function(yearsold, Linf, vbk, tknot) {
  pred_tl <- Linf * (1 - exp(-vbk * (yearsold - tknot)))
  return(pred_tl)
}

get_likelihood <- function(yearsold, tl, Linf, vbk, tknot, sig) {
  pred_tl <- predict_length(yearsold, Linf, vbk, tknot)
  NLL <- -1 * sum(dnorm(tl, pred_tl, sig, log=T), na.rm=T)
  return(NLL)
}
```

```{r}
map_columns <- function(v, cols) {
    c <- rep(0, 8)
    i <- 0
    for (col in cols) {
        i <- i + 1
        if (endsWith(col, 'Linf')) {
            j <- 1
        } else if (endsWith(col, 'vbk')) {
            j <- 3
        } else if (endsWith(col, 'tknot')) {
            j <- 5
        } else {
            j <- 7
        }

        if (startsWith(col, 'f_')) {
            c[j] <- v[i]
        } else if (startsWith(col, 'm_')) {
            c[j+1] <- v[i]
        } else {
            c[j] <- v[i]
            c[j+1] <- v[i]
        }
    }
    return(c)
}

map_columns(c(1, 3, 2, 3, 4), c('m_Linf', 'f_Linf', 'vbk', 'tknot', 'sig'))
```

```{r}
male_data <- filtered_data[filtered_data$Sex == 'M',]
female_data <- filtered_data[filtered_data$Sex == 'F',]
```

```{r}
do_likelihood_fit <- function(v, cols, runs) {
  objective <- function(v) {
    c <- map_columns(v, cols)

    f_Linf <- exp(c[1])
    f_vbk <- c[3]
    f_tknot <- c[5]
    f_sig <- exp(c[7])

    m_Linf <- exp(c[2])
    m_vbk <- c[4]
    m_tknot <- c[6]
    m_sig <- exp(c[8])

    f_NLL <- get_likelihood(female_data$Age, female_data$TL, f_Linf, f_vbk, f_tknot, f_sig)
    m_NLL <- get_likelihood(male_data$Age, male_data$TL, m_Linf, m_vbk, m_tknot, m_sig)

    NLL <- f_NLL + m_NLL
    return(NLL)
  }

  for (i in 1:runs) {
    fit <- optim(v, objective, hessian=T)
    v <- fit$par
  }
  return(fit)
}

summarize_aic <- function(fit, cols) {
  dof <- length(cols)
  nll <- fit$value
  aic <- 2*nll + 2*dof
  row <- c(paste(cols, collapse=','), dof, nll, aic)
  return(row)
}
```

## a. Use AIC model comparison to determine if growth rates are sexually dimorphic.

### All Parameters Free

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2,  4.1
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'f_sig',
  'm_Linf', 'm_vbk', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_free <- summarize_aic(fit, cols))
```

```{r}
plot(
    female_data$Age, 
    female_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship",
    col="orange"
)
points(
    male_data$Age, 
    male_data$TL, 
    xlab="Age", 
    ylab="Length", 
    col="blue"
)
lines(
    female_data$Age[order(female_data$Age)], 
    predict_length(
        female_data$Age[order(female_data$Age)], 
        exp(fit$par[1]), 
        fit$par[2], 
        fit$par[3]
    ), 
    col="orange"
)
lines(
    male_data$Age[order(male_data$Age)], 
    predict_length(
        male_data$Age[order(male_data$Age)], 
        exp(fit$par[5]), 
        fit$par[6], 
        fit$par[7]
    ), 
    col="blue"
)
legend(x=1.5, y=1300, pch=19, legend=c("Female", "Male"), col=c("orange", "blue"))
```

### One Parameter Shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  0.2, -1.2,  4.1
) 
cols <- c(
  'Linf', 'f_vbk', 'f_tknot', 'f_sig',
  'm_vbk', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(Linf_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  -1.2,  4.1
) 
cols <- c(
  'f_Linf', 'vbk', 'f_tknot', 'f_sig',
  'm_Linf', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(vbk_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, 4.1
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'f_sig',
  'm_Linf', 'm_vbk', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(tknot_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'sig',
  'm_Linf', 'm_vbk', 'm_tknot'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_shared <- summarize_aic(fit, cols))
```

### All Parameters Shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7
) 
cols <- c(
  'Linf', 'vbk', 'tknot', 'sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_shared <- summarize_aic(fit, cols))
```

### $\sigma$ and $t_0$ shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'sig',
  'm_Linf', 'm_vbk'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_and_tknot_shared <- summarize_aic(fit, cols))
```

## b. Report an AIC table.

```{r}
aic_table <- data.frame(rbind(
    all_free,
    all_shared,
    Linf_shared,
    vbk_shared,
    tknot_shared,
    sig_shared,
    sig_and_tknot_shared
))
colnames(aic_table) <- c('cols', 'dof', 'nll', 'aic')

aic_table$aic <- as.numeric(aic_table$aic)
aic_table$delta <- aic_table$aic - min(aic_table$aic)
(aic_table <- aic_table[order(aic_table$delta),])
```

These are definitely sexually dimorphic growth rates. 

## c. Which VBGF parameters differ between sexes in your top model?

The model with only $\sigma$ shared has the lowest AIC. Therefore $L_{\infty}$, $t_0$, and $K$ all differ between sexes.

## d. Report the VBGF parameter means and 95% confidence intervals from your top model
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'sig',
  'm_Linf', 'm_vbk', 'm_tknot'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_shared <- summarize_aic(fit, cols))
```

```{r}
stderr <- sqrt(diag(solve(fit$hessian)))
stderr

ALPHA <- 0.05
P_L95 <- (fit$par - qnorm(1-(ALPHA/2)) * stderr)
P_MEAN <- fit$par
P_U95 <- (fit$par + qnorm(1-(ALPHA/2)) * stderr)
P_U95[1] <- exp(P_U95[1])
P_MEAN[1] <- exp(P_MEAN[1])
P_L95[1] <- exp(P_L95[1])
P_U95[4] <- exp(P_U95[4])
P_MEAN[4] <- exp(P_MEAN[4])
P_L95[4] <- exp(P_L95[4])
P_U95[5] <- exp(P_U95[5])
P_MEAN[5] <- exp(P_MEAN[5])
P_L95[5] <- exp(P_L95[5])
```

```{r}
tab <- rbind(P_L95, P_MEAN, P_U95)
colnames(tab) <- c('f_Linf', 'f_K', 'f_t0', 'sigma', 'm_Linf', 'm_K', 'm_t0')
tab
```

## e. Plot the resulting male and female predicted length at age from your top model against the data.

```{r}
plot(
    female_data$Age, 
    female_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship",
    col="orange"
)
points(
    male_data$Age, 
    male_data$TL, 
    xlab="Age", 
    ylab="Length", 
    col="blue"
)
lines(
    female_data$Age[order(female_data$Age)], 
    predict_length(
        female_data$Age[order(female_data$Age)], 
        exp(fit$par[1]), 
        fit$par[2], 
        fit$par[3]
    ), 
    col="orange"
)
lines(
    male_data$Age[order(male_data$Age)], 
    predict_length(
        male_data$Age[order(male_data$Age)], 
        exp(fit$par[5]), 
        fit$par[6], 
        fit$par[7]
    ), 
    col="blue"
)
legend(x=1.5, y=1300, pch=19, legend=c("Female", "Male"), col=c("orange", "blue"))
```

# 3. Fit a series of logistic maturity models to the King Mackerel data. Determine if males and females exhibit sexually dimorphic maturity.

```{r}
filtered_data <- na.omit(data[,c('TL', 'Sex', 'Mat')])
female_data <- filtered_data[filtered_data['Sex'] == 'F',]
male_data <- filtered_data[filtered_data['Sex'] == 'M',]

get_nll <- function(theta) {
    f_lmat50 <- theta[1]
    f_sig <- theta[2]
    m_lmat50 <- theta[3]
    m_sig <- theta[4]

    prob_mat_female <- 1 / (1 + exp(-(female_data$TL - f_lmat50) / f_sig))
    prob_mat_male <- 1 / (1 + exp(-(male_data$TL - m_lmat50) / m_sig))
    nll_female <- -1*sum(dbinom(female_data$Mat, size=1, prob=prob_mat_female, log=T))
    nll_male <- -1*sum(dbinom(male_data$Mat, size=1, prob=prob_mat_male, log=T))
    return(nll_female + nll_male)
}
```

```{r}
subset_theta <- function(theta, params_to_fit) {
    input <- c()
    for (i in 1:length(params_to_fit)) {
        input <- append(input, theta[params_to_fit[i]])
    }
    return(input)
}

update_theta <- function(theta, input, params_to_fit, index_to_share) {
    v = theta
    for (i in 1:length(params_to_fit)) {
        v[params_to_fit[i]] <- input[i]
    }
    for (i in 1:length(v)) {
        if (!(i %in% params_to_fit)) {
            v[i] <- v[index_to_share[i]]
        }
    }
    return(v)
}

do_fit <- function(theta, params_to_fit, index_to_share) {

    fun <- function(input) {
        v = update_theta(theta, input, params_to_fit, index_to_share)
        return(get_nll(v))
    }

    input <- subset_theta(theta, params_to_fit)
    for (i in 1:10) {
        fit <- optim(input, fun, hessian=T)
        input <- fit$par
    }
    theta <- update_theta(theta, input, params_to_fit, index_to_share)
    return(theta)
}

get_aic <- function(theta, params_to_fit) {
    nll <- get_nll(theta)
    k <- length(params_to_fit)
    aic <- 2*k + 2*nll
    return(c(nll, k, aic))
}
```

## a. Use AIC model comparison to determine if maturity is different between sexes.

```{r}

starting_guess <- c(200, 20, 200, 20)
index_to_share <- c(3, 4, 1, 2)
row_names <- c()
col_names <- c("f_lmat50", "f_sig", "m_lmat50", "m_sig", "nll", "params", "AIC")

params_to_fit <- c(1, 2, 3)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row1 <- c(theta, aic_info)
row_names <- append(row_names, "L(h)sig(.)")

params_to_fit <- c(1, 2, 4)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row2 <- c(theta, aic_info)
row_names <- append(row_names, "L(.)sig(h)")

params_to_fit <- c(1, 2)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row3 <- c(theta, aic_info)
row_names <- append(row_names, "L(.)sig(.)")

params_to_fit <- c(1, 2, 3, 4)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row4 <- c(theta, aic_info)
row_names <- append(row_names, "L(h)sig(h)")
```

## b. Report an AIC table.
```{r}
results <- rbind(row1, row2, row3, row4)
colnames(results) <- col_names
rownames(results) <- row_names
results <- data.frame(results)

results <- results[order(results$AIC),]
results$delta_aic <- results$AIC - results$AIC[1]
results
```

## c. Which maturity parameters differ between sexes in your top model?
Technically the top model differs only in $\sigma$ but honestly the top two models are barely distuinguishable on the basis of AIC alone (I remember in class a difference of less than 5 being considered insignificant and this is much les than that). I'm going to go, in this case, with the model with the most degrees of freedom where both $L_{mat}$ and $\sigma$ are free. That way if there are biological differences we are capturing them. 

## d. Report the logistic maturity model parameter means and 95% confidence intervals from your top model.

```{r}
theta <- results[2, 1:4]
for (i in 1:10) {
    fit <- optim(theta, get_nll, hessian=T)
    theta <- fit$par
}
theta
```

```{r}
stderr <- sqrt(diag(solve(fit$hessian)))
stderr

ALPHA <- 0.05
P_L95 <- (fit$par - qnorm(1-(ALPHA/2)) * stderr)
P_MEAN <- fit$par
P_U95 <- (fit$par + qnorm(1-(ALPHA/2)) * stderr)

tab <- rbind(P_L95, P_MEAN, P_U95)
colnames(tab) <- c('f_lmat50', 'f_sig', 'm_lmat50', 'm_sig')
tab
```

## e. Plot the resulting male and female predicted maturity at length from your top model against the data.

```{r}
plot(
    female_data$TL,
    female_data$Mat,
    xlab="Length",
    ylab="Maturity",
    main="Female Maturity-Length Relationship",
)
lines(
    female_data$TL[order(female_data$TL)],
    1 / (1 + exp(-(female_data$TL[order(female_data$TL)] - fit$par[1]) / fit$par[2])),
    col="orange"
)
```

```{r}
plot(
    male_data$TL,
    male_data$Mat,
    xlab="Length",
    ylab="Maturity",
    main="Male Maturity-Length Relationship",
)
lines(
    male_data$TL[order(male_data$TL)],
    1 / (1 + exp(-(male_data$TL[order(male_data$TL)] - fit$par[3]) / fit$par[4])),
    col="blue"
)
```

# 4. Using the age composition data from the fishery dependent samples:

```{r}
fishery_data <- data[data$Fishery,]
aged <- na.omit(fishery_data[,c('Age', 'TL')])
aged$cmgrp <- floor(aged$TL / 10)
(length(aged$TL))
(length(fishery_data$TL))
(max(aged$Age))
head(aged)
```

## a. Generate an age-length key and project the age of samples with only lengths.

```{r}
get_ALK <- function(aged) {
    ALK <- data.frame(
        prop.table(
            table(aged$Age, aged$cmgrp),
            margin=2
        )
    )
    names(ALK) <- c("Age","cmgrp","prop") 
    ALK$Age <- as.numeric(as.character(ALK$Age))
    ALK$cmgrp <- as.numeric(as.character(ALK$cmgrp))
    ALK$prop <- as.numeric(as.character(ALK$prop))
    ALK$col.prop <- hcl.colors(101,'viridis')[round(ALK$prop,2)*100+1]
    return(ALK)
}

ALK <- get_ALK(aged)
head(ALK)
```

```{r}
get_catch_count_info <- function(catch, ALK) {
    catch_summary <- data.frame(table(catch$cmgrp))
    names(catch_summary) <- c("cmgrp", "total")
    catch_summary$cmgrp <- as.numeric(as.character(catch_summary$cmgrp))
    catch_summary$total <- as.numeric(as.character(catch_summary$total))
    info <- merge(ALK, catch_summary)
    info$count <- info$total * info$prop
    return(info)
}

catch <- na.omit(fishery_data[, c('TL', 'Fishery')])
catch$cmgrp <- floor(catch$TL / 10)
catch_info <- get_catch_count_info(catch, ALK)
head(catch_info)
```

```{r}
get_CAA <- function(catch_info) {
    CAA <- aggregate(count~Age, data=catch_info, sum)
    return(CAA)
}

(CAA <- get_CAA(catch_info))
```

### i. Plot a bubble plot of the age-length key.


```{r}
plot_ALK <- function(ALK, main) {
    # I want the circles to have area equal to
    # my proportion (for easier interpretation)
    # this would mean prop <- 4 * pi * r^2 or 
    # r <- sqrt(prop / (4 * pi))
    with(
        ALK, {
            symbols(
                x=cmgrp,
                y=Age,
                circles=sqrt(prop/(4 * pi)) * 2,
                inches=F,
                fg = col.prop,
                bg = col2rgbA(col.prop,0.5),
                xlab="Centimeter Group",
                ylab="Age",
                main=main
            )
        }
    )
}

plot_ALK(ALK, "Age-Length Key")
```

### ii. Plot the total catch at age data.

```{r}
with(
    CAA, {
        plot(
            Age,
            count,
            type='b',
            pch=16,
            col="black",
            xlab="Age",
            ylab="Count",
            main='Catch at Age'
        )
    }
)
```

## b. Estimate and report the instantaneous (Z) and finite (A) total mortality for this population. Justify the method you used to estimate Z and the data you included in the estimate.

We'll use the Millar model as the random effects allowed for in the model mean we can find the mortality pattern even with variability in the underlying recruitment for each age cohort. Furthermore we're going to only look at the years
for which the count per age has already peaked (in order to make sure we've got full recruitment to vulnerability). 

```{r}
reshape_for_millar <- function(CAA) {
    age_at_peak <- CAA$Age[which.max(CAA$count)]
    CAA_lim <- CAA[CAA$Age >= age_at_peak,]

    max_age <- max(CAA_lim$Age)
    CAA_ext <- rbind(
        CAA_lim,
        cbind(
            Age=(max_age+1):(2*max_age),
            count=rep(0,max_age)
        )
    )
    CAA_ext$count <- floor(CAA_ext$count)
    return(CAA_ext)
}

get_millar_fit <- function(CAA) {
    CAA_ext <- reshape_for_millar(CAA)

    fit <- glmer(
        count ~ Age + (1|Age),
        family=poisson,
        data=CAA_ext
    )
    print(summary(fit))
    return(fit)
}

(millar_fit <- get_millar_fit(CAA))
```

```{r}
get_params_from_millar <- function(millar_fit, alpha) {
    coef <- summary(millar_fit)$coef
    Z <- coef[2, 1]
    std <- coef[2, 2]
    ZU <- min(Z + qnorm(1-alpha/2)*std, 0)
    ZL <- Z - qnorm(1-alpha/2)*std

    A <- 1 - exp(Z)
    AU <- 1 - exp(ZU)
    AL <- 1 - exp(ZL)

    S <- exp(Z)
    SU <- exp(ZU)
    SL <- exp(ZL)

    m <- matrix(c(ZL, Z, ZU, AL, A, AU, SL, S, SU),3,3)
        colnames(m) <- c("Z", "A", "S")
        rownames(m) <- c("Lower", "MLE", "Upper")
    return(m)
}

get_params_from_millar(millar_fit, 0.05)
```

## c. Choose a single natural mortality surrogate method and estimate the natural mortality rate (M) for this population using your chosen surrogate method (use female-based parameters where appropriate).

Using the surrogate method from our Yield per Recruit analyses in the previous assignments we have:

$$M = -1.5K$$ 

```{r}
(M = -1.5 * 0.2257593)
```


## d. What is the fishing mortality rate (F) for this population assuming the M you estimated using surrogate methods?

```{r}
Z = -0.4718949
(F = Z - M)
```

### i. Discuss your level of confidence for this F estimate.

First of all we have to consider the uncertainty in Z itself. Z ranged from -0.51 to -0.43 which is a range of -0.08.
This range is itself equivalent to the value of F that we are estimated at present. Furthermore this M was taken from a
surrogate method which we know is highly approximate in the first place. So I'd definitely don't have a lot of confidence
in this estimate of F.

## e. What specific data would you collect to obtain a potentially better estimate of Z,M,and/or F?

I think we have a reasonable estimate of Z here. The real trick is being able to disentangle F and M. M can be really hard to estimate
on its own whereas F is more or less a matter of collecting the right data. Therefore I think it would be interesting to 
get an abundance estimate of the population in a given year along with a catch estimate in order to get a better sense of F 
and therefore M. 

# 5. Given the estimates of length-weight, growth, maturity, and mortality you estimated above (use your top models):

## a. Conduct an equilibrium Yield-per-Recruit analysis based solely on female King Mackerel.

### Length at Age
```{r}
A_max <- 23
ages <- seq(1, A_max, 1)

L_inf <- 1225.6
K <- 0.216
t_0 <- -2.1

get_lengths <- function(ages, L_inf, K, t_0){
  return(L_inf * (1 - exp(-K * (ages - t_0))))
}

lengths <- get_lengths(ages, L_inf, K, t_0)

plot(
  x=ages,
  y=lengths,
  type="l",
  xlab="Age",
  ylab="Length (mm)"
)
```

### Weight at Length
```{r}
a <- 7.01 * 10 ^ -9
b <- 3.01

get_weights <- function(lengths, a, b){
  return(a * lengths ^ b)
}

weights <- get_weights(lengths, a, b)

plot(
  x=lengths,
  y=weights,
  type="l",
  xlab="Length (mm)",
  ylab="Weight (kg)"
)
```


### Fecundity at Weight
```{r}
L_mat_50 <- 571.5

get_fecundity <- function(weights, L_mat_50) {
  W_mat_50 <- get_weights(L_mat_50, a, b)
  return(pmax(0, weights - W_mat_50))
}

fecundity <- get_fecundity(weights, L_mat_50)

plot(
  x=weights,
  y=fecundity,
  type="l",
  xlab="Weight (kg)",
  ylab="Fecundity"
)
```

### Vulnerability to Capture
```{r}
length_sigma <- 0.1
L_cap_50 <- 508

get_vul_cap <- function(lengths, L_cap_50, length_sigma) {
  sigma <- L_cap_50 * length_sigma
  return(
    1 / (1 + exp((L_cap_50 - lengths)/sigma))
  )
}

vul_cap <- get_vul_cap(lengths, L_cap_50, length_sigma)

plot(
  x=lengths,
  y=vul_cap,
  type="l",
  xlab="Length (mm)",
  ylab="Vulnerability to Capture"
)
```

### Chance Harvestable
```{r}
minLL <- 609.6
maxLL <- 2400

get_chance_harvestable <- function(lengths, minLL, maxLL, length_sigma) {
  max_sigma <- maxLL * length_sigma
  min_sigma <- minLL * length_sigma
  return(
    1 / (1 + exp((minLL - lengths)/min_sigma))
    - 1 / (1 + exp((maxLL - lengths)/max_sigma))
  )
}

chance_harvestable <- get_chance_harvestable(
  lengths, minLL, maxLL, length_sigma
)

plot(
  x=lengths,
  y=chance_harvestable,
  type="l",
  xlab="Length (mm)",
  ylab="Chance Harvestable"
)
```

### Survival Unfished
```{r}
get_S <- function(M) {
  return(exp(M))
}

get_natural_factor <- function(A_max, S) {
    return(c(1, rep(S, A_max - 1)))
}

M <- -0.17
natural_factor <- get_natural_factor(A_max, get_S(M))

plot(
  x=ages,
  y=cumprod(natural_factor),
  type="l",
  xlab="Age",
  ylab="Survival Unfished"
)
```

### Survival Fished
```{r}
U_harv <- 0.26
U_caught <- 0.0
D <- 0.2

get_harvest_factor <- function(
  U_harv, vul_cap, chance_harvestable
) {
  return(
    c(
      1, 1 - U_harv * chance_harvestable[1:length(chance_harvestable)-1]
    )
  )
}

get_discard_factor <- function(
    U_harv, U_caught, vul_cap, chance_harvestable, D
) {
  U <- U_harv + U_caught
  return(
    c(
      1, 1 - (
        U * vul_cap[1:length(vul_cap)-1] 
        - U_harv * chance_harvestable[1:length(chance_harvestable)-1]
      ) * D
    )
  )
}

harvest_factor <- get_harvest_factor(
  U_harv, vul_cap, chance_harvestable
)

discard_factor <- get_discard_factor(
  U_harv, U_caught, vul_cap, chance_harvestable, D
)

plot(
  x=ages,
  y=cumprod(natural_factor * harvest_factor * discard_factor),
  type="l",
  xlab="Age",
  ylab="Survival Fished"
)
```

## b. Report in a table all the parameters you used to conduct this analysis.

| Parameter | Value |
| --- | --- |
| $a_{max}$ | 23 |
| $a$ | 7.01e-09|
| $b$ | 3.01 | 
| $L_{\infty}$ |  1225.6 |
| $K$ | 0.216 |
| $t_0$ | -2.1 |
| $L_{mat}$ | 571.5 | 
| $M$ | -0.17|
| $F$ | -0.302 | 
| $U_{harv}$ | 0.26 |
| $U_{caught}$ | 0.0 | 
| $D$ | 0.2 |
| $L_{cap}$ | 508 |
| $minLL$ | 609.6 |
| $maxLL$ | 2400 | 

## c. Report in a table the life history, vulnerability, and survival vectors you used to conduct this analysis.

```{r}
build_table <- function(
  A_max, # ages to consider
  L_inf, K, t_0, # length params
  a, b, # weight params
  L_mat_50, # fecundity params
  L_cap_50, # vulnerability params
  minLL, maxLL, # harvestability params
  M, U_harv, U_caught, D, # survival params
  length_sigma # variability params
) {
  # build all of the modeling components
  ages <- seq(1, A_max, 1)
  lengths <- get_lengths(ages, L_inf, K, t_0)
  weights <- get_weights(lengths, a, b)
  fecundity <- get_fecundity(weights, L_mat_50)
  vul_cap <- get_vul_cap(lengths, L_cap_50, length_sigma)
  chance_harvestable <- get_chance_harvestable(
    lengths, minLL, maxLL, length_sigma
  )
  S <- get_S(M)
  natural_factor <- get_natural_factor(A_max, S)
  harvest_factor <- get_harvest_factor(
    U_harv, vul_cap, chance_harvestable
  )
  discard_factor <- get_discard_factor(
    U_harv, U_caught, vul_cap, chance_harvestable, D
  )
  tab <- rbind(
    lengths, weights, vul_cap, chance_harvestable, cumprod(natural_factor),
    cumprod(natural_factor * harvest_factor * discard_factor), fecundity
  )
  colnames(tab) <- ages
  rownames(tab) <- c(
    "length", "weight", "vul_cap", "vul_harvest", "survival_unfished",
    "survival_fished", "fecundity"
  )
  return(tab)
}

tab <- build_table(
  A_max, L_inf, K, t_0, a, b, L_mat_50, L_cap_50, minLL, maxLL, M, U_harv,
  U_caught, D, length_sigma
)
round(tab, 2)
```

## d. Do you believe that this King Mackerel population is undergoing either growth or recruitment overfishing?

```{r}
get_metrics <- function(tab, U_harv) {
    results <- c(
        sum(tab["survival_fished",] * tab["fecundity",]) / sum(tab["survival_unfished",] * tab["fecundity",]),
        sum(tab['weight',] * tab['vul_harvest',] * tab['survival_fished',]) * U_harv
    )
    return(results)
}

get_metrics(tab, U_harv)
```

Given SPR is 0.21 and the target is 0.35 we are definitely undergoing recruitment overfishing.

```{r}
U_vec <- seq(0, 1, 0.01)
yield_vec <- c()
for (U in U_vec) {
    tab <- build_table(
        A_max, L_inf, K, t_0, a, b, L_mat_50, L_cap_50, minLL, maxLL, M, U,
        U_caught, D, length_sigma
    )
    metrics <- get_metrics(tab, U)
    yield_vec <- c(yield_vec, metrics[2])
}

(U_vec[which.max(yield_vec)])
```

```{r}
plot(
    x=U_vec,
    y=yield_vec,
    type="l",
    xlab="U",
    ylab="Yield"
)
```

However it seems our yield is maximized at U = 0.27 which is just slightly higher than our U = 0.26. So growth overfishing is not occurring.

## e. What would be the minimum length limit necessary to maintain high yields and prevent growth and recruitment overfishing given the U?

```{r}
minLL_vec <- seq(0, 1500, 25)
yield_at_U_vec <- c()
U_at_max_yield_vec <- c()
spr_at_U_vec <- c()
U_vec <- seq(0, 1, 0.01)
for (new_minLL in minLL_vec) {
    yield_vec <- c()
    spr_vec <- c()
    for (U in U_vec) {
        tab <- build_table(
            A_max, L_inf, K, t_0, a, b, L_mat_50, L_cap_50, new_minLL, maxLL, M, U,
            U_caught, D, length_sigma
        )
        metrics <- get_metrics(tab, U)
        yield_vec <- c(yield_vec, metrics[2])
        spr_vec <- c(spr_vec, metrics[1])
    }
    yield_at_U_vec <- c(yield_at_U_vec, yield_vec[which(U_vec == U_harv)])
    spr_at_U_vec <- c(spr_at_U_vec, spr_vec[which(U_vec == U_harv)])
    U_at_max_yield_vec <- c(U_at_max_yield_vec, U_vec[which.max(yield_vec)])
}
```

```{r}
tab <- rbind(
    yield_at_U_vec, U_at_max_yield_vec, spr_at_U_vec
)
colnames(tab) <- minLL_vec
rownames(tab) <- c(
    "yield_at_U", "U_at_max_yield", "spr_at_U"
)
round(tab, 2)
```

In order to prevent growth and recruitment fishing we can simply filter down our data:

```{r}
t_tab <- data.frame(t(tab))
rownames(t_tab) <- minLL_vec
t_tab <- t_tab[t_tab$spr_at_U >= 0.35, ]
t_tab <- t_tab[t_tab$U_at_max_yield >= 0.26, ]
t_tab
```

Our highest yield is acheived at a slot limit of 875mm. Therefore the minimum slot limit required to maintain high yields while ensuring
we have neither growth or recruitment overfishing is 875mm.




