---
title: FAS6337C - Final Exam
author: Marcel Gietzmann-Sanders
---

```{r}
setwd("/workspaces/schooling/population_dynamics/final/") 
data <- read.table("data/King_Mackerel_sim-1.csv", header=T, sep=",") 
data$Age <- as.numeric(data$Age)
head(data)
```

# 1. Fit a length-weight relationship to the pooled (regardless of sex) data.

```{r}
data$lTL <- log(data$TL)
data$lWt <- log(data$Wt)
```

```{r}
plot(
    data$lTL, 
    data$lWt, 
    xlab="log(Length)", 
    ylab="log(Weight)", 
    main="Length-Weight Relationship"
)
```

## a. Report the length-weight parameters

```{r}
(fit <- lm(lWt ~ lTL, data=data))
```

```{r}
a <- exp(coef(fit)[1])
b <- coef(fit)[2]
c(a, b)
```

## b. Plot the mean weight at age against the data
```{r}
data$predWt <- a * data$TL^b
plot(
    data$TL, 
    data$Wt, 
    xlab="Length", 
    ylab="Weight", 
    main="Length-Weight Relationship"
)
lines(data$TL[order(data$TL)], data$predWt[order(data$TL)], col="red")
```

# 2. Fit a series of von Bertalanffy growth models to the King Mackerel data. Determine if males and females exihibit secually dimorphic growth.

```{r}
filtered_data <- na.omit(data)
head(filtered_data)
```

```{r}
plot(
    filtered_data$Age, 
    filtered_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship"
)
```

```{r}
predict_length <- function(yearsold, Linf, vbk, tknot) {
  pred_tl <- Linf * (1 - exp(-vbk * (yearsold - tknot)))
  return(pred_tl)
}

get_likelihood <- function(yearsold, tl, Linf, vbk, tknot, sig) {
  pred_tl <- predict_length(yearsold, Linf, vbk, tknot)
  NLL <- -1 * sum(dnorm(tl, pred_tl, sig, log=T), na.rm=T)
  return(NLL)
}
```

```{r}
map_columns <- function(v, cols) {
    c <- rep(0, 8)
    i <- 0
    for (col in cols) {
        i <- i + 1
        if (endsWith(col, 'Linf')) {
            j <- 1
        } else if (endsWith(col, 'vbk')) {
            j <- 3
        } else if (endsWith(col, 'tknot')) {
            j <- 5
        } else {
            j <- 7
        }

        if (startsWith(col, 'f_')) {
            c[j] <- v[i]
        } else if (startsWith(col, 'm_')) {
            c[j+1] <- v[i]
        } else {
            c[j] <- v[i]
            c[j+1] <- v[i]
        }
    }
    return(c)
}

map_columns(c(1, 3, 2, 3, 4), c('m_Linf', 'f_Linf', 'vbk', 'tknot', 'sig'))
```

```{r}
male_data <- filtered_data[filtered_data$Sex == 'M',]
female_data <- filtered_data[filtered_data$Sex == 'F',]
```

```{r}
do_likelihood_fit <- function(v, cols, runs) {
  objective <- function(v) {
    c <- map_columns(v, cols)

    f_Linf <- exp(c[1])
    f_vbk <- c[3]
    f_tknot <- c[5]
    f_sig <- exp(c[7])

    m_Linf <- exp(c[2])
    m_vbk <- c[4]
    m_tknot <- c[6]
    m_sig <- exp(c[8])

    f_NLL <- get_likelihood(female_data$Age, female_data$TL, f_Linf, f_vbk, f_tknot, f_sig)
    m_NLL <- get_likelihood(male_data$Age, male_data$TL, m_Linf, m_vbk, m_tknot, m_sig)

    NLL <- f_NLL + m_NLL
    return(NLL)
  }

  for (i in 1:runs) {
    fit <- optim(v, objective, hessian=T)
    v <- fit$par
  }
  return(fit)
}

summarize_aic <- function(fit, cols) {
  dof <- length(cols)
  nll <- fit$value
  aic <- 2*nll + 2*dof
  row <- c(paste(cols, collapse=','), dof, nll, aic)
  return(row)
}
```

## a. Use AIC model comparison to determine if growth rates are sexually dimorphic.

### All Parameters Free

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2,  4.1
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'f_sig',
  'm_Linf', 'm_vbk', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_free <- summarize_aic(fit, cols))
```

```{r}
plot(
    female_data$Age, 
    female_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship",
    col="orange"
)
points(
    male_data$Age, 
    male_data$TL, 
    xlab="Age", 
    ylab="Length", 
    col="blue"
)
lines(
    female_data$Age[order(female_data$Age)], 
    predict_length(
        female_data$Age[order(female_data$Age)], 
        exp(fit$par[1]), 
        fit$par[2], 
        fit$par[3]
    ), 
    col="orange"
)
lines(
    male_data$Age[order(male_data$Age)], 
    predict_length(
        male_data$Age[order(male_data$Age)], 
        exp(fit$par[5]), 
        fit$par[6], 
        fit$par[7]
    ), 
    col="blue"
)
```

### One Parameter Shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  0.2, -1.2,  4.1
) 
cols <- c(
  'Linf', 'f_vbk', 'f_tknot', 'f_sig',
  'm_vbk', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(Linf_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  -1.2,  4.1
) 
cols <- c(
  'f_Linf', 'vbk', 'f_tknot', 'f_sig',
  'm_Linf', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(vbk_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, 4.1
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'f_sig',
  'm_Linf', 'm_vbk', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(tknot_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'sig',
  'm_Linf', 'm_vbk', 'm_tknot'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_shared <- summarize_aic(fit, cols))
```

### All Parameters Shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7
) 
cols <- c(
  'Linf', 'vbk', 'tknot', 'sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_shared <- summarize_aic(fit, cols))
```

### $\sigma$ and $t_0$ shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'sig',
  'm_Linf', 'm_vbk'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_and_tknot_shared <- summarize_aic(fit, cols))
```

## b. Report an AIC table.

```{r}
aic_table <- data.frame(rbind(
    all_free,
    all_shared,
    Linf_shared,
    vbk_shared,
    tknot_shared,
    sig_shared,
    sig_and_tknot_shared
))
colnames(aic_table) <- c('cols', 'dof', 'nll', 'aic')

aic_table$aic <- as.numeric(aic_table$aic)
aic_table$delta <- aic_table$aic - min(aic_table$aic)
(aic_table <- aic_table[order(aic_table$delta),])
```

These are definitely sexually dimorphic growth rates. The model with all parameters free has the lowest AIC.

## c. Which VBGF parameters differ between sexes in your top model?

Given $\sigma$ and $t_0$ being shared is within 4 AIC of the fully free model I'm going to say that $L_{\infty}$ and $K$ are different between sexes. 

## d. Report the VBGF parameter means and 95% confidence intervals from your top model
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'sig',
  'm_Linf', 'm_vbk'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_and_tknot_shared <- summarize_aic(fit, cols))
```

```{r}
stderr <- sqrt(diag(solve(fit$hessian)))
stderr

ALPHA <- 0.05
P_L95 <- (fit$par - qnorm(1-(ALPHA/2)) * stderr)
P_MEAN <- fit$par
P_U95 <- (fit$par + qnorm(1-(ALPHA/2)) * stderr)
P_U95[1] <- exp(P_U95[1])
P_MEAN[1] <- exp(P_MEAN[1])
P_L95[1] <- exp(P_L95[1])
P_U95[4] <- exp(P_U95[4])
P_MEAN[4] <- exp(P_MEAN[4])
P_L95[4] <- exp(P_L95[4])
P_U95[5] <- exp(P_U95[5])
P_MEAN[5] <- exp(P_MEAN[5])
P_L95[5] <- exp(P_L95[5])
```

```{r}
tab <- rbind(P_L95, P_MEAN, P_U95)
colnames(tab) <- c('f_Linf', 'f_K', 't0', 'sigma', 'm_Linf', 'm_K')
tab
```

## e. Plot the resulting male and female predicted length at age from your top model against the data.

```{r}
plot(
    female_data$Age, 
    female_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship",
    col="orange"
)
points(
    male_data$Age, 
    male_data$TL, 
    xlab="Age", 
    ylab="Length", 
    col="blue"
)
lines(
    female_data$Age[order(female_data$Age)], 
    predict_length(
        female_data$Age[order(female_data$Age)], 
        exp(fit$par[1]), 
        fit$par[2], 
        fit$par[3]
    ), 
    col="orange"
)
lines(
    male_data$Age[order(male_data$Age)], 
    predict_length(
        male_data$Age[order(male_data$Age)], 
        exp(fit$par[5]), 
        fit$par[6], 
        fit$par[3]
    ), 
    col="blue"
)
legend(x=1.5, y=1300, pch=19, legend=c("Female", "Male"), col=c("orange", "blue"))
```







