---
title: FAS6337C - Final Exam
author: Marcel Gietzmann-Sanders
---

```{r}
setwd("/workspaces/schooling/population_dynamics/final/") 
data <- read.table("data/King_Mackerel_sim-1.csv", header=T, sep=",") 
data$Age <- as.numeric(data$Age)
head(data)
```

# 1. Fit a length-weight relationship to the pooled (regardless of sex) data.

```{r}
data$lTL <- log(data$TL)
data$lWt <- log(data$Wt)
```

```{r}
plot(
    data$lTL, 
    data$lWt, 
    xlab="log(Length)", 
    ylab="log(Weight)", 
    main="Length-Weight Relationship"
)
```

## a. Report the length-weight parameters

```{r}
(fit <- lm(lWt ~ lTL, data=data))
```

```{r}
a <- exp(coef(fit)[1])
b <- coef(fit)[2]
c(a, b)
```

## b. Plot the mean weight at age against the data
```{r}
data$predWt <- a * data$TL^b
plot(
    data$TL, 
    data$Wt, 
    xlab="Length", 
    ylab="Weight", 
    main="Length-Weight Relationship"
)
lines(data$TL[order(data$TL)], data$predWt[order(data$TL)], col="red")
```

# 2. Fit a series of von Bertalanffy growth models to the King Mackerel data. Determine if males and females exihibit secually dimorphic growth.

```{r}
filtered_data <- na.omit(data)
head(filtered_data)
```

```{r}
plot(
    filtered_data$Age, 
    filtered_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship"
)
```

```{r}
predict_length <- function(yearsold, Linf, vbk, tknot) {
  pred_tl <- Linf * (1 - exp(-vbk * (yearsold - tknot)))
  return(pred_tl)
}

get_likelihood <- function(yearsold, tl, Linf, vbk, tknot, sig) {
  pred_tl <- predict_length(yearsold, Linf, vbk, tknot)
  NLL <- -1 * sum(dnorm(tl, pred_tl, sig, log=T), na.rm=T)
  return(NLL)
}
```

```{r}
map_columns <- function(v, cols) {
    c <- rep(0, 8)
    i <- 0
    for (col in cols) {
        i <- i + 1
        if (endsWith(col, 'Linf')) {
            j <- 1
        } else if (endsWith(col, 'vbk')) {
            j <- 3
        } else if (endsWith(col, 'tknot')) {
            j <- 5
        } else {
            j <- 7
        }

        if (startsWith(col, 'f_')) {
            c[j] <- v[i]
        } else if (startsWith(col, 'm_')) {
            c[j+1] <- v[i]
        } else {
            c[j] <- v[i]
            c[j+1] <- v[i]
        }
    }
    return(c)
}

map_columns(c(1, 3, 2, 3, 4), c('m_Linf', 'f_Linf', 'vbk', 'tknot', 'sig'))
```

```{r}
male_data <- filtered_data[filtered_data$Sex == 'M',]
female_data <- filtered_data[filtered_data$Sex == 'F',]
```

```{r}
do_likelihood_fit <- function(v, cols, runs) {
  objective <- function(v) {
    c <- map_columns(v, cols)

    f_Linf <- exp(c[1])
    f_vbk <- c[3]
    f_tknot <- c[5]
    f_sig <- exp(c[7])

    m_Linf <- exp(c[2])
    m_vbk <- c[4]
    m_tknot <- c[6]
    m_sig <- exp(c[8])

    f_NLL <- get_likelihood(female_data$Age, female_data$TL, f_Linf, f_vbk, f_tknot, f_sig)
    m_NLL <- get_likelihood(male_data$Age, male_data$TL, m_Linf, m_vbk, m_tknot, m_sig)

    NLL <- f_NLL + m_NLL
    return(NLL)
  }

  for (i in 1:runs) {
    fit <- optim(v, objective, hessian=T)
    v <- fit$par
  }
  return(fit)
}

summarize_aic <- function(fit, cols) {
  dof <- length(cols)
  nll <- fit$value
  aic <- 2*nll + 2*dof
  row <- c(paste(cols, collapse=','), dof, nll, aic)
  return(row)
}
```

## a. Use AIC model comparison to determine if growth rates are sexually dimorphic.

### All Parameters Free

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2,  4.1
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'f_sig',
  'm_Linf', 'm_vbk', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_free <- summarize_aic(fit, cols))
```

```{r}
plot(
    female_data$Age, 
    female_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship",
    col="orange"
)
points(
    male_data$Age, 
    male_data$TL, 
    xlab="Age", 
    ylab="Length", 
    col="blue"
)
lines(
    female_data$Age[order(female_data$Age)], 
    predict_length(
        female_data$Age[order(female_data$Age)], 
        exp(fit$par[1]), 
        fit$par[2], 
        fit$par[3]
    ), 
    col="orange"
)
lines(
    male_data$Age[order(male_data$Age)], 
    predict_length(
        male_data$Age[order(male_data$Age)], 
        exp(fit$par[5]), 
        fit$par[6], 
        fit$par[7]
    ), 
    col="blue"
)
```

### One Parameter Shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  0.2, -1.2,  4.1
) 
cols <- c(
  'Linf', 'f_vbk', 'f_tknot', 'f_sig',
  'm_vbk', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(Linf_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  -1.2,  4.1
) 
cols <- c(
  'f_Linf', 'vbk', 'f_tknot', 'f_sig',
  'm_Linf', 'm_tknot', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(vbk_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, 4.1
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'f_sig',
  'm_Linf', 'm_vbk', 'm_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(tknot_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'f_tknot', 'sig',
  'm_Linf', 'm_vbk', 'm_tknot'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_shared <- summarize_aic(fit, cols))
```

### All Parameters Shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7
) 
cols <- c(
  'Linf', 'vbk', 'tknot', 'sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_shared <- summarize_aic(fit, cols))
```

### $\sigma$ and $t_0$ shared
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'sig',
  'm_Linf', 'm_vbk'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_and_tknot_shared <- summarize_aic(fit, cols))
```

## b. Report an AIC table.

```{r}
aic_table <- data.frame(rbind(
    all_free,
    all_shared,
    Linf_shared,
    vbk_shared,
    tknot_shared,
    sig_shared,
    sig_and_tknot_shared
))
colnames(aic_table) <- c('cols', 'dof', 'nll', 'aic')

aic_table$aic <- as.numeric(aic_table$aic)
aic_table$delta <- aic_table$aic - min(aic_table$aic)
(aic_table <- aic_table[order(aic_table$delta),])
```

These are definitely sexually dimorphic growth rates. The model with all parameters free has the lowest AIC.

## c. Which VBGF parameters differ between sexes in your top model?

Given $\sigma$ and $t_0$ being shared is within 4 AIC of the fully free model I'm going to say that $L_{\infty}$ and $K$ are different between sexes. 

## d. Report the VBGF parameter means and 95% confidence intervals from your top model
```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2
) 
cols <- c(
  'f_Linf', 'f_vbk', 'tknot', 'sig',
  'm_Linf', 'm_vbk'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_and_tknot_shared <- summarize_aic(fit, cols))
```

```{r}
stderr <- sqrt(diag(solve(fit$hessian)))
stderr

ALPHA <- 0.05
P_L95 <- (fit$par - qnorm(1-(ALPHA/2)) * stderr)
P_MEAN <- fit$par
P_U95 <- (fit$par + qnorm(1-(ALPHA/2)) * stderr)
P_U95[1] <- exp(P_U95[1])
P_MEAN[1] <- exp(P_MEAN[1])
P_L95[1] <- exp(P_L95[1])
P_U95[4] <- exp(P_U95[4])
P_MEAN[4] <- exp(P_MEAN[4])
P_L95[4] <- exp(P_L95[4])
P_U95[5] <- exp(P_U95[5])
P_MEAN[5] <- exp(P_MEAN[5])
P_L95[5] <- exp(P_L95[5])
```

```{r}
tab <- rbind(P_L95, P_MEAN, P_U95)
colnames(tab) <- c('f_Linf', 'f_K', 't0', 'sigma', 'm_Linf', 'm_K')
tab
```

## e. Plot the resulting male and female predicted length at age from your top model against the data.

```{r}
plot(
    female_data$Age, 
    female_data$TL, 
    xlab="Age", 
    ylab="Length", 
    main="Length-Age Relationship",
    col="orange"
)
points(
    male_data$Age, 
    male_data$TL, 
    xlab="Age", 
    ylab="Length", 
    col="blue"
)
lines(
    female_data$Age[order(female_data$Age)], 
    predict_length(
        female_data$Age[order(female_data$Age)], 
        exp(fit$par[1]), 
        fit$par[2], 
        fit$par[3]
    ), 
    col="orange"
)
lines(
    male_data$Age[order(male_data$Age)], 
    predict_length(
        male_data$Age[order(male_data$Age)], 
        exp(fit$par[5]), 
        fit$par[6], 
        fit$par[3]
    ), 
    col="blue"
)
legend(x=1.5, y=1300, pch=19, legend=c("Female", "Male"), col=c("orange", "blue"))
```

# 3. Fit a series of logistic maturity models to the King Mackerel data. Determine if males and females exhibit sexually dimorphic maturity.

```{r}
get_nll <- function(theta) {
    f_lmat50 <- theta[1]
    f_sig <- theta[2]
    m_lmat50 <- theta[3]
    m_sig <- theta[4]

    prob_mat_female <- 1 / (1 + exp(-(female_data$TL - f_lmat50) / f_sig))
    prob_mat_male <- 1 / (1 + exp(-(male_data$TL - m_lmat50) / m_sig))
    nll_female <- -1*sum(dbinom(female_data$Mat, size=1, prob=prob_mat_female, log=T))
    nll_male <- -1*sum(dbinom(male_data$Mat, size=1, prob=prob_mat_male, log=T))
    return(nll_female + nll_male)
}
```

```{r}
subset_theta <- function(theta, params_to_fit) {
    input <- c()
    for (i in 1:length(params_to_fit)) {
        input <- append(input, theta[params_to_fit[i]])
    }
    return(input)
}

update_theta <- function(theta, input, params_to_fit, index_to_share) {
    v = theta
    for (i in 1:length(params_to_fit)) {
        v[params_to_fit[i]] <- input[i]
    }
    for (i in 1:length(v)) {
        if (!(i %in% params_to_fit)) {
            v[i] <- v[index_to_share[i]]
        }
    }
    return(v)
}

do_fit <- function(theta, params_to_fit, index_to_share) {

    fun <- function(input) {
        v = update_theta(theta, input, params_to_fit, index_to_share)
        return(get_nll(v))
    }

    input <- subset_theta(theta, params_to_fit)
    for (i in 1:10) {
        fit <- optim(input, fun, hessian=T)
        input <- fit$par
    }
    theta <- update_theta(theta, input, params_to_fit, index_to_share)
    return(theta)
}

get_aic <- function(theta, params_to_fit) {
    nll <- get_nll(theta)
    k <- length(params_to_fit)
    aic <- 2*k + 2*nll
    return(c(nll, k, aic))
}
```

## a. Use AIC model comparison to determine if growth rates are sexually dimorphic.

```{r}

starting_guess <- c(200, 20, 200, 20)
index_to_share <- c(3, 4, 1, 2)
row_names <- c()
col_names <- c("f_lmat50", "f_sig", "m_lmat50", "m_sig", "nll", "params", "AIC")

params_to_fit <- c(1, 2, 3)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row1 <- c(theta, aic_info)
row_names <- append(row_names, "L(h)sig(.)")

params_to_fit <- c(1, 2, 4)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row2 <- c(theta, aic_info)
row_names <- append(row_names, "L(.)sig(h)")

params_to_fit <- c(1, 2)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row3 <- c(theta, aic_info)
row_names <- append(row_names, "L(.)sig(.)")

params_to_fit <- c(1, 2, 3, 4)
theta <- do_fit(starting_guess, params_to_fit, index_to_share)
aic_info <- get_aic(theta, params_to_fit)
row4 <- c(theta, aic_info)
row_names <- append(row_names, "L(h)sig(h)")
```

## b. Report an AIC table.
```{r}
results <- rbind(row1, row2, row3, row4)
colnames(results) <- col_names
rownames(results) <- row_names
results <- data.frame(results)

results <- results[order(results$AIC),]
results$delta_aic <- results$AIC - results$AIC[1]
results
```

## c. Which maturity parameters differ between sexes in your top model?
Technically the top model differs only in $\sigma$ but honestly the top three models are barely distuinguishable on the basis of AIC alone. I'm going to go, in this case, with the model with the most degrees of freedom where both $L_{mat}$ and $\sigma$ are free. That way if there are biological differences we are capturing them. 

## d. Report the logistic maturity model parameter means and 95% confidence intervals from your top model.

```{r}
theta <- results[2, 1:4]
for (i in 1:10) {
    fit <- optim(theta, get_nll, hessian=T)
    theta <- fit$par
}
theta
```

```{r}
stderr <- sqrt(diag(solve(fit$hessian)))
stderr

ALPHA <- 0.05
P_L95 <- (fit$par - qnorm(1-(ALPHA/2)) * stderr)
P_MEAN <- fit$par
P_U95 <- (fit$par + qnorm(1-(ALPHA/2)) * stderr)

tab <- rbind(P_L95, P_MEAN, P_U95)
colnames(tab) <- c('f_lmat50', 'f_sig', 'm_lmat50', 'm_sig')
tab
```

## e. Plot the resulting male and female predicted maturity at length from your top model against the data.

```{r}
plot(
    female_data$TL,
    female_data$Mat,
    xlab="Length",
    ylab="Maturity",
    main="Female Maturity-Length Relationship",
)
lines(
    female_data$TL[order(female_data$TL)],
    1 / (1 + exp(-(female_data$TL[order(female_data$TL)] - fit$par[1]) / fit$par[2])),
    col="orange"
)
```

```{r}
plot(
    male_data$TL,
    male_data$Mat,
    xlab="Length",
    ylab="Maturity",
    main="Male Maturity-Length Relationship",
)
lines(
    male_data$TL[order(male_data$TL)],
    1 / (1 + exp(-(male_data$TL[order(male_data$TL)] - fit$par[3]) / fit$par[4])),
    col="blue"
)
```







