---
title: STAT641 - Homework 1
author: Marcel Gietzmann-Sanders
---

# 1. General Questions

a. Marcel Gietzmann-Sanders - 31308957
b. 
c. No 
d. Just wanting a broad introduction to the subject.
e. I'm studying fisheries and frequentists stats is everywhere. I'd like to get a better handle on the Bayesian side of things. 
f. Yes 
g. This will be the first time.
h. Yes 
i. I've been a software engineer working primarily in Python for the past 7 years. All of my classes here have used R. So at this point I probably have over 10,000 hours of Python and several hundred hours of R experience. 
j. I'm on a macbook but I typically run a Linux VM for my work. 
k. No

# 2. Opening Problems

### (a) Use R to simulate the following; state the values it generates.

```{r}
rbinom(5, size=12, prob=0.4)
```

$n$ is the number of trials per experiment and $p$ is the probability of a success (or equivalently, of a 1) per trial.

```{r}
rnorm(5, mean=0.0, sd=0.75)
```

### (b) Confidence Intervals

> I purchase a random sample of n = 49 avocados at Fred Meyer. The sample mean
>weight is 4.3 ounces with a standard deviation of 0.36 ounces. Calculate a 95%
>confidence interval for the mean weight of all such avocados, and interpret your
>confidence interval. (Be sure to state the formula you’re using, and fill in the various
>values in the formula.) What, exactly, is it that happens with 95% probability?
>What does the “95%” refer to when constructing a 95% confidence interval.

The formula for a 95% confidence interval is:

$$\bar{\mu} \pm Z_{0.975}\frac{s}{\sqrt{n}}$$

where $\bar{\mu}$ is the sample mean, $s$ is the sample standard deviation, and $Z_{0.975}$ is the z-score at which the cumulative probability of a normal distribution is 97.5%. 
Because this is two tailed, choosing the 97.5% ensures that we are covering 95% of the distribution. 

In our case this becomes:

$$4.3\pm 1.96 \frac{0.36}{\sqrt{49}}=4.3\pm 0.1008$$

So this is saying (informally) that we are 95% confident that the true mean weight of avocados at Fred Meyer 
lies somewhere between 4.1992 and 4.4008 ounces. Formally if we were to repeat this experiment over and over
95% of the time the true mean weight would lie within the confidence interval derived in that experiment. 

### (c) Hypothesis Testing

> I conducted a poll consisting of a random sample of n = 1000 sane individuals. 790
> of those polled answered “Yes” to the question, “Is the world going insane?” Test
> the hypotheses, H0 : p = 0.75 versus Ha : p > 0.75, where p is the proportion of all
> sane individuals who believe the world is going insane. (It’s okay to conduct the test
> by hand or to use the R function prop.test.) At level α = 0.05, do we reject H0 or
> fail to reject H0?
>
> It turns out that the p-value for this test is 0.00174. At level α = 0.01, should we
> reject H0? Why or why not?


```{r}
prop.test(790, 1000, p=0.75, alternative="greater", correct=FALSE)
```

In this case the p-value is 0.00174 which is far below our $\alpha=0.05$ threshold. Therefore we can go ahead and reject the null hypothesis. 

We should also reject the null hypothesis given an $\alpha=0.01$ because we are still below the threshold. 
