---
title: STAT641 - Homework 2
author: Marcel Gietzmann-Sanders
---

## Problem 1

(a) I suppose I would wonder whether this measurement is all that extreme given
my initial hypothesis. More precisely I'd wonder what the probability finding at 
least this many hypertensive people in a selection of five would be given my
hypothesis. 

```{r}
sum(dbinom(c(4, 5), 5, 0.3))
```

Given this is exceedingly small (only ~3% likely given my hypothesis)
I would likely reject my initial hypothesis and decide that $\theta$ is likely 
higher than 0.3. I wouldn't necessarily say $\theta=0.2$ given the small sample size
but it's certainly higher than 0.3. 

(b) At this point, given the large sample size I would likely estimate $\theta=\frac{400}{1000}=0.4$.
I have enough data here that whatever ill conceived notions I had at the beginning had better be 
laid to rest. 

## Problem 2

Let's start by gathering what we know. 

$$P(Y=1|\theta = 1)=0.975$$
$$P(Y=0|\theta=0)=0.95$$
$$P(\theta=1)=0.01$$

(a) What we are concerned with here is the probability that someone has the
disease given they tested positive - $P(\theta=1|Y=1)$.

Using Bayes rule we have:

$$P(\theta=1|Y=1)=\frac{P(Y=1|\theta=1)P(\theta=1)}{P(Y=1)}$$

We know $P(Y=1|\theta=1)$ and $P(\theta=1)$ so all that remains is to determine 
$P(Y=1)$

$$P(Y=1)=P(Y=1|\theta=0)P(\theta=0)+P(Y=1|\theta=1)P(\theta=1)$$

$$P(Y=1)=\left[1-P(Y=0|\theta=0)\right]\left[1-P(\theta=1)\right]+P(Y=1|\theta=1)P(\theta=1)$$

$$P(Y=1)=0.05(0.99) + 0.975(0.01)=0.05925$$

Therefore we have:

$$P(\theta=1|Y=1)=\frac{0.975(0.01)}{0.05925}\approx0.165$$

So of those who test postive for the disease approximately 16.5% of them 
will actually have the disease. 

(b) In this case we're interested in $P(\theta=0|Y=0)$

$$P(\theta=0|Y=0)=\frac{P(\theta=0|Y=0)P(\theta=0)}{P(Y=0)}$$

$$P(\theta=0|Y=0)=\frac{P(\theta=0|Y=0)\left[1-P(\theta=1)\right]}{\left[1-P(Y=0)\right]}$$

$$P(\theta=0|Y=0)=\frac{0.95(0.99)}{0.94075}\approx 0.9997$$

So if someone takes the test and gets a negative there is a approximately 99.97% chance that
they indeed are disease free. If you get a negative there's more or less no need to be worried.

## Problem 3

Let $L$ stand in for our color of labrador with 1, 2, 3 being yellow, 
chocolate, and black respectively. Let $B=1$ mean they love to play 
fetch and $B=0$ mean that they don't. We are interested in the following:

$$P(L=x|B=1)$$

In general:

$$P(L=x|B=1)=\frac{P(B=1|L=x)P(L=x)}{P(B=1)}$$

We have the numerators for our various colors so all we really need to 
determine is $P(B=1)$. But this is just:

$$P(B=1)=\sum_x P(B=1|L=x)P(L=x)$$

$$P(B=1)=0.95(0.5)+0.92(0.3) +0.9(0.2)=0.931$$

So now we have:

$$P(L=1|B=1)=\frac{0.95(0.5)}{0.931}\approx 0.51$$

$$P(L=2|B=1)=\frac{0.92(0.3)}{0.931}\approx 0.3$$

$$P(L=3|B=1)=\frac{0.9(0.2)}{0.931}\approx 0.19$$

So if we know the dog likes to play fetch there's a 51% probability it's a yellow,
30% that its a chocolate lab, and 19% that its a black lab. Clearly playing fetch 
is not particularly discriminating. 

## Problem 4

a. In this case the likelihood is just:

$$p(3|\theta)= {8\choose 3}\theta^3(1-\theta)^5=56\theta^3(1-\theta)^5$$

b. Let's plot it!

```{r}
n <- 8
y <- 3
theta <- seq(0, 1, length=200)
likelihood <- choose(n,y) * theta ^ y * (1 - theta) ^ (n - y)
plot(theta, likelihood, type='l')
```
