---
title: STAT641 - Homework 6
author: Marcel Gietzmann-Sanders
---

```{r}
rejection_sampling = function(M, p, dg, rg, N) {
    samples = c()
    num_samples = 0
    num_rejected = 0
    while (length(samples) < N) {
        num_samples = num_samples + 1
        # sample a theta from g
        theta = rg()
        # flip a coin
        R = p(theta) / (dg(theta) * M)
        success = rbinom(1, 1, R)
        if (success == 1) {
            samples = append(samples, theta)
        } else {
            num_rejected = num_rejected + 1
        }
    }
    return(list(samples = samples, num_samples = num_samples, num_rejected=num_rejected))
}
```

## Problem 1

### (a)

```{r, fig.height = 2.5, fig.width = 6, fig.align = "center"}
theta = seq(0, 1, 0.01)
q = 0.05 + (theta - 0.5) ^ 2
par(mar = c(4,4,1,1))
plot(x=theta, y=q, type="l")
```

### (b)

We want 

$$p(\theta)=kq(\theta)$$

s.t. 

$$\int_0^1 p(\theta)d\theta = 1$$

This means we want:

$$k=\left(\int_0^1 q(\theta)d\theta \right)^{-1}$$

$$\int q(\theta)d\theta=\int (0.05 + (\theta-0.5)^2)d\theta=0.05\theta + \frac{1}{3}(\theta-0.5)^3 + C$$

$$\int_0^1 q(\theta)d\theta = 0.05(1) + \frac{1}{3}(1-0.5)^3 - 0.05(0) - \frac{1}{3}(0-0.5)^3=\frac{1}{20}+\frac{2}{3}\left(\frac{1}{2}\right)^3$$
$$=\frac{1}{20}+\frac{1}{12}=\frac{2}{15}$$

Therefore:

$$k=\frac{15}{2}$$

```{r, fig.height = 2.5, fig.width = 6, fig.align = "center"}
theta = seq(0, 1, 0.01)
func_p = function(theta) {
    return((0.05 + (theta - 0.5) ^ 2) * 15 / 2)
}
p = func_p(theta)
par(mar = c(4,4,1,1))
plot(x=theta, y=p,type="l")
```

### (c) 

A $Beta(1,1)$ distribution is just the uniform distribution over $[0,1]$. This means that

$$g(\theta)\sim Beta(1, 1)$$

$$g(\theta) = 1, \theta \in [0, 1] $$

Therefore for:

$$\frac{p(\theta)}{g(\theta)}\leq M$$

we require that $M=\max{p(\theta)}$. Given our $p(\theta)$ is centered at 0.5 and is a 
upward parabola the maximums will be at the edges of $\theta$'s range. I.e.

$$M=\frac{15}{2}\left( 0.05 + (0-0.5)^2\right)=\frac{9}{4}$$

```{r, fig.height = 3, fig.width = 6, fig.align = "center"}
M = 9/4
dg = function(theta) {
    return(1)
}
rg = function() {
    return(rbeta(1, 1, 1))
}

results = rejection_sampling(M, func_p, dg, rg, 10000)
par(mar = c(4,4,1,1))
hist(results$samples, prob=TRUE, xlab="theta", main="")
lines(x=theta, y=p, col="red")
```


\newpage 

The fraction of samples thrown away is just over 55%. 

```{r}
results$num_rejected / results$num_samples
```

### (d)

The mean is $\approx 0.5$ and the variance is $\approx 0.125$.

```{r}
(mean(results$samples))
(var(results$samples))
```

### (e)

In this case $g\sim Beta(0.5,0.5)$ which presents us with a much trickier problem 
for finding $M$. 

```{r, fig.height = 2.5, fig.width = 6, fig.align = "center"}
par(mar = c(4,4,1,1))
plot(x=theta, y=dbeta(theta, 0.5, 0.5), type="l")
```

What we require is that:

$$\frac{p(\theta)}{g(\theta)}\leq M$$

\newpage

So graphing this ratio we find:

```{r, fig.height = 2.5, fig.width = 6, fig.align = "center"}
par(mar = c(4,4,1,1))
ratio = func_p(theta) / dbeta(theta, 0.5, 0.5)
plot(x=theta,y=ratio,type="l")
```

This ratio has a clear maximum value so let's go ahead and zoom in on one of those peaks.

```{r, fig.height = 2.5, fig.width = 6, fig.align = "center"}
zoom_theta = seq(0, 0.2, 0.0001)
zoom_ratio = func_p(zoom_theta) / dbeta(zoom_theta, 0.5, 0.5)
par(mar = c(4,4,1,1))
plot(x=zoom_theta,y=zoom_ratio,type="l", xlab="theta", ylab="ratio")
```

The maximum here is simply:

```{r}
max(ratio)
```

$M=1.5$ should do! 

\newpage

```{r}
M = 1.5
dg = function(theta) {
    return(dbeta(theta, 0.5, 0.5))
}
rg = function() {
    return(rbeta(1, 0.5, 0.5))
}

results = rejection_sampling(M, func_p, dg, rg, 10000)
```

```{r, fig.height = 3, fig.width = 6, fig.align = "center"}
par(mar = c(4,4,1,1))
hist(results$samples, prob=TRUE, xlab="theta", main="")
lines(x=theta, y=p, col="red")
```

The fraction of samples thrown away is just over 33%. 

```{r}
results$num_rejected / results$num_samples
```

And once again the mean is $\approx 0.5$ and the variance is $\approx 0.125$.

```{r}
(mean(results$samples))
(var(results$samples))
```

### (f)

I'd say the second blanketing function $g(\theta)\sim Beta(0.5, 0.5)$ was better
just because we reject far fewer samples and therefore require less compute to get
the same number of accepted samples. 

## Problem 2

### (a)

$$q(\theta) = |\sin(\theta)|e^{-\theta/4}, \theta\geq 0$$

```{r, fig.height = 3, fig.width = 6, fig.align = "center"}
par(mar = c(4,4,1,1))
func_q = function(theta) {
    abs(sin(theta)) * exp(-theta / 4)
}
theta = seq(0, 25, 0.01)
q = func_q(theta)
plot(x=theta,y=q,type="l")
```

### (b) 

We'll be using an exponential distribution as our blanketing function. I.e. 

$$g(\theta) = \lambda e^{-\lambda \theta}$$

Now we can think of our $q$ as being a $\sin$ function 
multiplied by an exponential envelope. Therefore we want our $\lambda$ to give us this same 
envelope.  

$$g(\theta) = \frac{1}{4} e^{- \theta /4}$$

So now we have:

$$\frac{q(\theta)}{g(\theta)}=\frac{|\sin(\theta)|e^{-\theta/4}}{(1/4)e^{- \theta /4}}=4|\sin(\theta)|\leq M$$

Which clearly means we'll choose $M=4$ as the maximum magnitude of $\sin$ is just 1.
\newpage
```{r}
M = 4
dg = function(theta) {
    return(dexp(theta, 1/4))
}
rg = function() {
    return(rexp(1, 1/4))
}

results = rejection_sampling(M, func_q, dg, rg, 10000)
```

### (c)

```{r, fig.height = 3, fig.width = 6, fig.align = "center"}
theta = seq(min(results$samples), max(results$samples), 0.01)
q = func_q(theta)
par(mar=c(5, 4, 4, 6) + 0.1)
hist(results$samples, prob=TRUE, xlab="theta", main="", breaks=100)
par(new=TRUE)
plot(x=theta, y=q, type="l", axes=FALSE, col="red")
axis(4, col="red",col.axis="red",las=1)
```

The proportion of rejected samples is roughly 37%:

```{r}
results$num_rejected / results$num_samples
```


### (d) 

And here the mean is $\approx 4.1$ and the variance is $\approx 16$.

```{r}
(mean(results$samples))
(var(results$samples))
```
