---
title: FISH621 - Exam
author: Marcel Gietzmann-Sanders
---

```{r}
library(tidyverse)
```

# Problem 1: Blackfish

```{r}
data = read.csv("Twill Pond Blackfish.csv")
head(data)
```

### 1. Calculate a representative index of sampling effort for each day.

We'll go with total trap hours as our unit of effort.

```{r}
data$trap_hours = data$Traps * data$Hours.Fished
head(data)
```

### 2. Use a depletion estimator to determine the catchability of blackfish in Twill Pond.

```{r}
data$cumulative_catch = cumsum(data$Catch) - data$Catch
data$CPUE = data$Catch / data$trap_hours
head(data)
```

```{r}
model = lm(data$CPUE ~ data$cumulative_catch)
```

```{r}
plot(data$CPUE ~ data$cumulative_catch)
points(data$cumulative_catch, predict(model), col="red")
legend("topright", legend=c("Observed", "Fit"), col=c("black", "red"), pch=1)
```


Given we are fitting:

$$CPUE = qN_0 - qK$$

We can extract the catchability coefficient $q$ from the model:

```{r}
(q = -model$coefficients[2])
```

### 3. Calculate an estimate of blackfish abundance in Twill Pond at the start of your experiment.

```{r}
(N = model$coefficients[1] / q)
```

# Problem 2: Mark-recapture

### 1. Calculate Petersen estimates of abundance for each of these five ponds.

```{r}
data = read.csv("mr.dat.csv")
head(data)
```

```{r}
data$petersen = data$n1 * data$n2 / data$m2
head(data)
```

### 2. Calculate Chapman estimates of abundance for each pond, along with the variance in these estimates and the coefficient of variation (CV) for each abundance estimate.

```{r}
data$chapman = (data$n1 + 1) * (data$n2 + 1) / (data$m2 + 1) - 1
data$chapman.var = (
    (data$n1 + 1) * (data$n2 + 1) * (data$n1 - data$m2) * (data$n2 - data$m2) 
    / ((data$m2 + 1)^2 * (data$m2 + 2))
)
data$chapman.cv = sqrt(data$chapman.var) / data$chapman
head(data[,c("Site", "chapman", "chapman.var", "chapman.cv")])
```

### 3. Calculate Bailey estimates of abundance for each pond, along with variances and associated CV's.

```{r}
data$bailey = data$n1 * (data$n2 + 1) / (data$m2 + 1)
data$bailey.var = (
    data$n1^2 * (data$n2 + 1) * (data$n2 - data$m2) 
    / ((data$m2 + 1)^2 * (data$m2 + 2))
)
data$bailey.cv = sqrt(data$chapman.var) / data$chapman
head(data[,c("Site", "bailey", "bailey.var", "bailey.cv")])
```

### 4. Please describe under what conditions of this experiment the Bailey as opposed to Petersen or Chapman estimates should be most appropriate.

Being a binominal model the Bailey estimator assumes that recaptures of individuals are possible whereas in Chapman this is not the case. Practically this means that
the probability of capture does not change as you sample individuals (because in theory you're putting those individuals back as you sample them). 
Therefore the Bailey estimator is most appropriate when either your sample size is very small (and therefore there are only minimal changes to the 
sample probability) or if you are in fact able to recapture individuals. But in general if recapture isn't possible you might as well as go with the Chapman
estimator. 

The other case where the probability of capture may not change dramatically (and certainly not as much as a hypergeometric distribution would expect)
is if you do something like stratified sampling where you aren't actually sampling the same population of individuals but instead are taking individuals 
from a series of different strata. Then the probability of capture in each strata is (in theory) not affected by what happens in other strata. So the 
Bailey model may be useful here as well.

# Problem 3: Pacific Cod

### 1. A design-based biomass index.

```{r}
trawl = read.csv("race_goa_pcod.csv")
stratum = read.csv("Stratum Descriptions.csv")
stratum = stratum[stratum$Survey == "GOA",]
head(trawl)
head(stratum)
```

```{r}
unique(trawl$Scientific.Name)
```

```{r}
cpue_by_stratum = trawl %>% group_by(Year, Stratum) %>% summarize(CPUE = mean(Weight.CPUE..kg.km2.))
area_by_stratum = stratum %>% select(Stratum, Area..km2.)
by_stratum = inner_join(cpue_by_stratum, area_by_stratum, by="Stratum")
by_stratum$biomass = by_stratum$CPUE * by_stratum$Area..km2.
total_biomass = by_stratum %>% group_by(Year) %>% summarize(biomass = sum(biomass))
plot(
    x=total_biomass$Year,
    y=total_biomass$biomass,
    xlab="Year",
    ylab="Total Biomass Estimate",
    type="l"
)
```

```{r}
cpue_by_stratum.var = trawl %>% group_by(Year, Stratum) %>% summarize(CPUE.var = var(Weight.CPUE..kg.km2.), stations = n())
cpue_by_stratum.var = cpue_by_stratum.var[!is.na(cpue_by_stratum.var$CPUE.var),] # sometimes we have single samples
by_stratum = inner_join(cpue_by_stratum.var, area_by_stratum, by="Stratum")
by_stratum$biomass.var = by_stratum$CPUE.var * (by_stratum$Area..km2. ^ 2) / by_stratum$stations
total_biomass.var = by_stratum %>% group_by(Year) %>% summarize(biomass.var = sum(biomass.var))
by_year = inner_join(total_biomass, total_biomass.var, by="Year")
plot(
    x=by_year$Year,
    y=by_year$biomass.var,
    xlab="Year",
    ylab="Estimate of Variance",
    type="l"
)
```

The table:

```{r}
by_year$biomass.sd = sqrt(by_year$biomass.var)
by_year$biomass.cv = by_year$biomass.sd / by_year$biomass
by_year
```

### 2. Model-based biomass index with a year effect only.

```{r}
trawl$fYear = factor(trawl$Year)
glm.year = glm(
    log(Weight.CPUE..kg.km2. + 1) ~ fYear, data=trawl
)
by_year = inner_join(by_year, trawl %>% distinct(Year, fYear), by="Year")
by_year$glm.year = exp(predict(glm.year, newdata=by_year)) - 1
by_year
```

### 3. Model-based biomass index that controls for among strata variation in average biomass.

```{r}
trawl$fStratum = factor(trawl$Stratum)
glm.year_stratum = glm(
    log(Weight.CPUE..kg.km2. + 1) ~ fYear + fStratum, data=trawl
)
by_year$fStratum = trawl$fStratum[1]
by_year$glm.year_stratum = exp(predict(glm.year_stratum, newdata=by_year)) - 1
by_year[,c("biomass", "glm.year", "glm.year_stratum")]
```

### 4. Finally, please plot the three biomass indices to compare and interpret their trends across time, in addition to your table of annual values.

```{r}
by_year$biomass.norm = by_year$biomass / max(by_year$biomass)
by_year$glm.year.norm = by_year$glm.year / max(by_year$glm.year)
by_year$glm.year_stratum.norm = by_year$glm.year_stratum / max(by_year$glm.year_stratum)
plot.default(
    by_year$Year, 
    by_year$biomass.norm, 
    type="l",
    col="orange", 
    xlab="Year", 
    ylab="Biomass Index", 
    main="Comparison of Biomass Indices"
)
lines(
    by_year$Year, 
    by_year$glm.year.norm, 
    col="cyan"
)
lines(
    by_year$Year, 
    by_year$glm.year_stratum.norm, 
    col="purple"
)
legend(
    "bottomleft", 
    legend=c("Design-based", "Model (Year)", "Model (Year + Stratum)"), 
    col=c("orange", "cyan", "purple"), 
    lty=1
)
```

```{r}
by_year[,c("Year", "biomass", "biomass.var", "biomass.sd", "biomass.cv", "glm.year", "glm.year_stratum")]
```

The model based indices seem to follow the same pattern reasonably well whereas the design based
index has a pretty extreme peak in 2009. It's interesting to note that the variance and the CV both 
are exceptionally high in 2009 as well. Add onto this the fact that the stratum based model is the
least "peaky" of the three and it seems likely that one of the strata in that year had highly variable
CPUE. 

It's also interesting to note that while the design based index has a peak around 1996 the model based 
indices just seem this as part of a downward trend. 

Finally the stratum and year based model in later years (past 2000) seems 
to be slightly under the year only model. So perhaps a "more productive" stratum is being over represented 
in the later as opposed to the earlier years.

# Problem 4: Wild Hogs


```{r}
data = data.frame(
    stratum = c(1, 2, 3, 4),
    sampling_units = c(200, 100, 500, 200),
    sampled_units = c(50, 60, 20, 10),
    abundance_per_unit = c(45, 56, 32, 20),
    variance_per_stratum = c(1200, 3000, 2900, 6000)
)
head(data)
```

### 1. An estimate of total hog abundance within each stratum.
```{r}
data$abundance_per_stratum = data$sampling_units * data$abundance_per_unit
data[,c("stratum", "abundance_per_stratum")]
```

### 2. The estimated hog population total in the entire region.
```{r}
(total_pop = sum(data$abundance_per_stratum))

```

### 3. An unbiased estimate of the variance of the hog population total.

```{r}
(total_pop.var = sum(
    data$sampling_units * (data$sampling_units - data$sampled_units) * data$variance_per_stratum / data$sampled_units
))
```

### 4. The coefficient of variation for the estimate of the hog population total.

```{r}
(total_pop.cv = sqrt(total_pop.var) / total_pop)
```

### 5. An approximate 95% confidence interval for the population total.

```{r}
alpha = 0.05
z = qnorm(1-alpha/2)
lower = total_pop - z * sqrt(total_pop.var)
upper = total_pop + z * sqrt(total_pop.var)
c(lower, upper)
```

### 6. An unbiased estimate of mean hog abundance per 1-km2 for the entire population.

```{r}
(pop_per_unit = total_pop / sum(data$sampling_units))
```

### 7. An unbiased estimate of the variance of the mean hog abundance per 1-km2.

```{r}
(pop_per_unit.var = sum(
    (data$sampling_units / sum(data$sampling_units)) ^ 2 * ((data$sampling_units - data$sampled_units) / sum(data$sampling_units)) * data$variance_per_stratum / data$sampled_units
))
```

### 8. An approximate 95% confidence interval for the mean hog abundance per 1-km2

```{r}
alpha = 0.05
z = qnorm(1-alpha/2)
lower = pop_per_unit - z * sqrt(pop_per_unit.var)
upper = pop_per_unit + z * sqrt(pop_per_unit.var)
c(lower, upper)
```

# Problem 5: Experimental Design

### 1. Movement among spatial units within the scope of your mark-recapture experiment. 

For this experiment we'd need to first choose our spatial units and then use distinct tags 
for each spatial unit. This would allow us to then track which fish have come from where. 

### 2. Survival among sampling periods within the mark-recapture experiment.

Here we could take advantage of a Jolly-Seber design. We'd repeatedly capture,
count marked individuals, mark unmarked individuals and release them. Then the Jolly-Seber 
model allows us to estimate the "survival" rate among sampling periods. Note though 
that "survival" will include both emigration and mortality. 

### 3. Differences across time in the probability of detection across sampling periods.

Here we could use individual mark recapture with a time model. This model assumes that 
the detection probability is the same across all individuals but is allowed to differ over
time (i.e. sampling periods).

# Thank You!
