---
title: FISH621 - Homework 3
author: Marcel Gietzmann-Sanders
---

```{r}
setwd("/workspaces/schooling/abundance/hwk_3/")
```

# Problem 1


```{r}
data = read.table("data/Schnabel Data.csv", header=T, sep=",")
head(data)
```

#### 1. Calculate the number of fish marked in each sampling period $u_i$

```{r}
data$u = data$Captures - data$Recaptures
data$u
```

#### 2. Calculate the total number of marked fish in the population at the start of each sampling period ($M_i$)

```{r}
data$M = cumsum(data$u) - data$u
data$M
```

#### 3. Determine the total number of trout in the lake using the basic Schnabel estimator.

```{r}
N.schnabel = sum(data$Captures * data$M)/sum(data$Recaptures)
N.schnabel
```


#### 4. Determine the total number of trout in the lake using the Chapman extension to the Schnabel estimator.

```{r}
N.chapman = sum(data$Captures * data$M)/(sum(data$Recaptures) + 1)
N.chapman
```

#### 5. Using the Chapman extension to the Schnabel estimator, please calculate an approximate 95% confidence interval for this estimate

```{r}
lambda = sum(data$Captures * data$M)
m = sum(data$Recaptures)

alpha = 0.05
z = qnorm(1-alpha/2)

upper = (
    lambda * (
        (2 * m + z ^ 2 + z * sqrt(4 * m + z ^ 2))
        / (2 * (m ^ 2))
    )
)
lower =  (
    lambda * (
        (2 * m + z ^ 2 - z * sqrt(4 * m + z ^ 2))
        / (2 * (m ^ 2))
    )
)
c(lower, upper)
```

#### 6. Calculate separate Chapman abundance estimates for each sampling period: 2: s.

```{r}
n = length(data$Captures)
data$N.chapman[2:n] = (data$M[2:n] + 1) * (data$Captures[2:n] + 1) / (data$Recaptures[2:n] + 1) - 1
data$N.chapman
```

#### 7. Calculate the variance estimate for each of the Chapman abundance estimates for sampling periods: 2:s.

```{r}
data$var.chapman = (
    (
        (data$M + 1) * (data$Captures + 1) 
        * (data$M - data$Recaptures) * (data$Captures - data$Recaptures)
    ) / (
        (data$Recaptures + 1) ^ 2 * (data$Recaptures + 2)
    )
)
data$var.chapman
```

#### 8. Find the mean Chapman estimator for total trout abundance from this experiment

```{r}
N.chapman.mean = mean(data$N.chapman[2:n])
N.chapman.mean
```

#### 9. Calculate the theoretical variance for the mean Chapman estimator, and approximate 95% confidence interval for this estimate.

```{r}
var.chapman.mean = sum(data$var.chapman[2:n]) / (n - 1)^2
var.chapman.mean
```

```{r}
z = qnorm(1-alpha/2)
lower = N.chapman.mean - z * sqrt(var.chapman.mean)
upper = N.chapman.mean + z * sqrt(var.chapman.mean)
c(lower, upper)
```

#### 10. Calculate the empirical variance for the mean Chapman estimator, and approximate 95% confidence interval for this estimate.

```{r}
var.chapman.mean.emp = sum((data$N.chapman[2:n] - N.chapman.mean)^2) / (n - 1) / (n - 2)
var.chapman.mean.emp
```

```{r}
z = qnorm(1-alpha/2)
lower = N.chapman.mean - z * sqrt(var.chapman.mean.emp)
upper = N.chapman.mean + z * sqrt(var.chapman.mean.emp)
c(lower, upper)
```

#### 11. Include a table describing $n_i$ , $m_i$ , $u_i$ , $M_i$ , Chapman estimates, and the coefficient of variation for Chapman estimates, across sampling periods 1:s.

```{r}
data$n = data$Captures
data$m = data$Recaptures
data$cv = sqrt(data$var.chapman) / data$N.chapman
data[, c("n", "m", "u", "M", "N.chapman", "cv")]
```

# Problem 2

#### 1. An estimate of total wolf abundance within each stratum: $\tau_h$.

```{r}
stratum = 1:3
n = c(50, 50, 50)
y = c(10, 20, 30)
var = c(2800, 700, 600)
size = c(100, 50, 300)
data = data.frame(stratum, n, y, var, size)
data
```

```{r}
data$tau = y * size 
data$tau
```

#### 2. The estimated wolf population total in the entire 450-km2 region: $\tau_{st}$

```{r}
total_pop = sum(data$tau)
total_pop
```

#### 3. An unbiased estimate of the variance of the wolf population total: $var({t_{st}})$

```{r}
total_pop.var = sum(data$size * (data$size - data$n) * data$var / data$n)
total_pop.var
```

#### 4. The coefficient of variation for the estimate of the wolf population total.

```{r}
sqrt(total_pop.var) / total_pop
```

#### 5. An approximate 95% confidence interval for the population total.

```{r}
z = qnorm(1-alpha/2)
lower = total_pop - z * sqrt(total_pop.var)
upper = total_pop + z * sqrt(total_pop.var)
c(lower, upper)
```

#### 6. An unbiased estimate of mean wolf abundance per 1-km2 for the entire population: $\mu_{st}$

```{r}
mu = total_pop / sum(data$size)
mu 
```

#### 7. An unbiased estimate of the variance of the mean wolf abundance per 1-km2

```{r}
var.mu = sum(
    (data$size / sum(data$size))^2 * (data$size - data$n) / (data$size) * data$var / data$n
)
var.mu
```

#### 8. An approximate 95% confidence interval for the mean wolf abundance per 1-km2 

```{r}
z = qnorm(1-alpha/2)
lower = mu - z * sqrt(var.mu)
upper = mu + z * sqrt(var.mu)
c(lower, upper)
```

# Problem 3
