---
title: FAS6337C - Lab 3
author: Marcel Gietzmann-Sanders
---

# Growth Models in R

Data for this laboratory are from two Florida Spotted Seatrout Cynoscion nebulosus populations (Indian River and Charlotte Harbor). The data set contains only females from each population. Fish were collected from 1986 to 1988 by Murphy and Taylor (1994, Transactions of the American Fisheries Society 123:482-497).

```{r}
setwd("/workspaces/schooling/population_dynamics/lab_3/")
trout_data <- read.table("data/trout.txt", header=T, sep="") 
head(trout_data)
```

The objectives of this laboratory are:

1. To determine the best growth model for each population and estimate $L_{\infty}$, $t_0$, and $k$;
2. Estimate the mean length at age for each population;
3. Use an AIC model selection method to compare growth curves and parameters between populations.

Please conduct the following analyses in R and answer the questions. Each question or bullet point should be responded to either with text, a table, or a plot. Use trout.txt as your data file.

## 1. Use nonlinear least squares (nls) to estimate the von Bertalanffy growth parameters separately for Charlotte Harbor and for Indian River populations. 

```{r}
ch_data <- na.omit(
    trout_data[trout_data$bay == 'CharlotteHarbor',]
)
ch_data <- ch_data[order(ch_data$yearsold),]
ir_data <- na.omit(
    trout_data[trout_data$bay == 'IndianRiver',]
)
ir_data <- ir_data[order(ir_data$yearsold),]
```

```{r}
do_least_squares_fit <- function(fish_data, Linf, vbk, tknot) {
  tl <- as.numeric(fish_data$tl)
  yearsold <- as.numeric(fish_data$yearsold)
  result <- nls(
    tl ~ Linf * (1 - exp(-vbk * (yearsold - tknot))),
    data=fish_data,
    start=list(Linf=Linf, vbk=vbk, tknot=tknot)
  )
  return(result)
}

ch_result = do_least_squares_fit(ch_data, 800, 0.2, -1.4)
ir_result = do_least_squares_fit(ir_data, 950, 0.19, -1.2)
```

```{r}
summary(ch_result)
```

```{r}
summary(ir_result)
```


```{r}
ch_data$nls_tl <- ch_result$m$predict(ch_data$yearsold)
ir_data$nls_tl <- ir_result$m$predict(ir_data$yearsold)

ch_data$nls_res <- ch_data$tl - ch_data$nls_tl
ir_data$nls_res <- ir_data$tl - ir_data$nls_tl

head(ch_data)
```

### Plot the residuals against the predicted values from the nls for each population. 

```{r}
with(
  ch_data, {
    plot(
      nls_res ~ nls_tl,
      main='Charlotte Harbor Residuals Plot',
      xlab='Predicted Length (mm)',
      ylab='Actual - Predicted Length (mm)'
    )
  }
)
```

```{r}
with(
  ir_data, {
    plot(
      nls_res ~ nls_tl,
      main='Indian River Residuals Plot',
      xlab='Predicted Length (mm)',
      ylab='Actual - Predicted Length (mm)'
    )
  }
)
```

```{r}
with(
  ch_data, {
    plot(
      tl ~ yearsold,
      main='Charlotte Harbor Fit',
      xlab='Years Old',
      ylab='Length (mm)'
    )
    lines(nls_tl ~ yearsold, col='red')
  }
)
```


```{r}
with(
  ir_data, {
    plot(
      tl ~ yearsold,
      main='Indian River Fit',
      xlab='Years Old',
      ylab='Length (mm)'
    )
    lines(nls_tl ~ yearsold, col='red')
  }
)
```

### Does it appear that a von Bertalanffy curve is reasonable for each population? 

The conclusion here is the same as in the last lab. The fit is reasonable but far from ideal. There is a clear pattern from under predicting to overpredicting as length increases so there's definitely a pattern to the residuals that (in theory) could be captured in a better model. 

There's also quite a lot of variance left in these estimates, and that variance itself is patterned with our gear selectivity creating variance toward smaller predicted lengths.

## 2. For the Charlotte Harbor population, create a function that estimates the negative log likelihood of the normal distribution. Minimize the negative log likelihood of the normal distribution to estimate the parameters of the von Bertalanffy growth equation. 


```{r}
predict_length <- function(yearsold, Linf, vbk, tknot) {
  pred_tl <- Linf * (1 - exp(-vbk * (yearsold - tknot)))
  return(pred_tl)
}

get_likelihood <- function(yearsold, tl, Linf, vbk, tknot, sig) {
  pred_tl <- predict_length(yearsold, Linf, vbk, tknot)
  NLL <- -1 * sum(dnorm(tl, pred_tl, sig, log=T), na.rm=T)
  return(NLL)
}

do_likelihood_fit <- function(fish_data, Linf, vbk, tknot, sig) {
  tl <- as.numeric(fish_data$tl)
  yearsold <- as.numeric(fish_data$yearsold)
  lLinf <- log(Linf)
  lsig <- log(sig)

  objective <- function(v) {
    Linf <- exp(v[1])
    vbk <- v[2]
    tknot <- v[3]
    sig <- exp(v[4])
    NLL <- get_likelihood(yearsold, tl, Linf, vbk, tknot, sig)
    return(NLL)
  }

  v <- c(lLinf, vbk, tknot, lsig)
  fit <- optim(v, objective, hessian=T)
  print("Fit Summary")
  print(fit)


  covm <- solve(fit$hessian)

  pred <- c(
    exp(fit$par[1]),
    fit$par[2],
    fit$par[3],
    exp(fit$par[4])
  )
  print(pred)
  return(list(pred, covm, fit$par))
}

ch_ml_result <- do_likelihood_fit(ch_data, 800, 0.2, -1.4, 40)
ch_pred_params <- ch_ml_result[[1]]
ch_covm <- ch_ml_result[[2]]
ch_ml_par <- ch_ml_result[[3]]

ch_data$ml_tl <- predict_length(
  ch_data$yearsold, ch_pred_params[1], ch_pred_params[2], ch_pred_params[3]
)
ch_data$ml_res <- ch_data$tl - ch_data$ml_tl
```

### Plot the residuals against the predicted values from the likelihood estimation

```{r}
with(
  ch_data, {
    plot(
      ml_res ~ ml_tl,
      main='Charlotte Harbor Residuals Plot',
      xlab='Predicted Length (mm)',
      ylab='Actual - Predicted Length (mm)'
    )
  }
)
```

```{r}
with(
  ch_data, {
    plot(
      tl ~ yearsold,
      main='Charlotte Harbor Fit',
      xlab='Years Old',
      ylab='Length (mm)'
    )
    lines(ml_tl ~ yearsold, col='red')
  }
)
```

### Does the likelihood estimation appear to return VBGM parameter estimates that are a reasonable model for the data? 

$t_0$ is clearly off and the model seems to be overpredicting $L_{\infty}$. Furthermore there is clearly changing variance with both age and length so the $\sigma$ parameter is also contentious. Overall the fit to the data given is reasonable. 

### Compare your estimates to the between the nls (question 1) and the likelihood estimation. 

Besides having added $\sigma$ this is the same fit. So the estimates are the same.

## 3. Generate the confidence intervals for each parameter from the model using the output from nls and the Hessian matrix from Optim. 

### Report the confidence intervals for each parameter from each method in a table

```{r}
ml_stderr <- sqrt(diag(ch_covm))

ALPHA <- 0.05
ML_L95 <- (ch_ml_par - qnorm(1-(ALPHA/2)) * ml_stderr)
ML_U95 <- (ch_ml_par + qnorm(1-(ALPHA/2)) * ml_stderr)
ML_U95[1] <- exp(ML_U95[1])
ML_L95[1] <- exp(ML_L95[1])
ML_U95[4] <- exp(ML_U95[4])
ML_L95[4] <- exp(ML_L95[4])
```



```{r}
ch_nls_par <- summary(ch_result)$parameters[,1]
ch_nls_stderr <- summary(ch_result)$parameters[,2]
NLS_L95 <- (ch_nls_par - qnorm(1-(ALPHA/2)) * ch_nls_stderr)
NLS_U95 <- (ch_nls_par + qnorm(1-(ALPHA/2)) * ch_nls_stderr)
```


```{r}
NLS_U95['sig'] <- NaN
NLS_L95['sig'] <- NaN
rbind(ML_U95, ML_L95, NLS_U95, NLS_L95)
```

### Compare the estimated confidence intervals between methods. 

$\sigma$ obviously offers no comparison as it is not in both methods. $t_0$ and $k$ are both very similar between the two methods. $L_{\infty}$ is the only one that stands out but it was fit as the log of itself during maximum likelihood estimation and so it's expected that the confidence interval for $L_{\infty}$ is asymmetric in the likelihood estimation.

## 4. Conduct the same analysis for the Indian River population (questions 1-3). Answer the same questions/bullet points. 


```{r}
ir_ml_result <- do_likelihood_fit(ir_data, 950, 0.19, -1.2, 40)
ir_pred_params <- ir_ml_result[[1]]
ir_covm <- ir_ml_result[[2]]
ir_ml_par <- ir_ml_result[[3]]

ir_data$ml_tl <- predict_length(
  ir_data$yearsold, ir_pred_params[1], ir_pred_params[2], ir_pred_params[3]
)
ir_data$ml_res <- ir_data$tl - ir_data$ml_tl
```

### Plot the residuals against the predicted values from the likelihood estimation

```{r}
with(
  ir_data, {
    plot(
      ml_res ~ ml_tl,
      main='Indian River Residuals Plot',
      xlab='Predicted Length (mm)',
      ylab='Actual - Predicted Length (mm)'
    )
  }
)
```

```{r}
with(
  ir_data, {
    plot(
      tl ~ yearsold,
      main='Indian River Fit',
      xlab='Years Old',
      ylab='Length (mm)'
    )
    lines(ml_tl ~ yearsold, col='red')
  }
)
```

### Does the likelihood estimation appear to return VBGM parameter estimates that are a reasonable model for the data? 

$t_0$ is clearly off again. There is still clearly changing variance with both age and length so the $\sigma$ parameter is also contentious. Overall the fit to the data given is reasonable. 

### Compare your estimates to the between the nls (question 1) and the likelihood estimation. 

Besides having added $\sigma$ this is the same fit. So the estimates are the same.

### Report the confidence intervals for each parameter from each method in a table

```{r}
ml_stderr <- sqrt(diag(ir_covm))

ALPHA <- 0.05
ML_L95 <- (ir_ml_par - qnorm(1-(ALPHA/2)) * ml_stderr)
ML_U95 <- (ir_ml_par + qnorm(1-(ALPHA/2)) * ml_stderr)
ML_U95[1] <- exp(ML_U95[1])
ML_L95[1] <- exp(ML_L95[1])
ML_U95[4] <- exp(ML_U95[4])
ML_L95[4] <- exp(ML_L95[4])
```



```{r}
ir_nls_par <- summary(ir_result)$parameters[,1]
ir_nls_stderr <- summary(ir_result)$parameters[,2]
NLS_L95 <- (ir_nls_par - qnorm(1-(ALPHA/2)) * ir_nls_stderr)
NLS_U95 <- (ir_nls_par + qnorm(1-(ALPHA/2)) * ir_nls_stderr)
```


```{r}
NLS_U95['sig'] <- NaN
NLS_L95['sig'] <- NaN
rbind(ML_U95, ML_L95, NLS_U95, NLS_L95)
```

### Compare the estimated confidence intervals between methods. 

It's interesting that here all parameters are extremely close in terms of bounds - even $L_{\infty}$. However the confidence interval for $L_{\infty}$ is still asymmetric for the maximum likelihood estimation. 

## 5. Using the likelihood-based model parameter estimates for each population, calculate the mean length at each age for age 0 to age 8 for each population. 

### Report the mean length at age for each population. 


```{r}
age <- seq(0, 8, 1)
ch_tl <- predict_length(
  age, ch_pred_params[1], ch_pred_params[2], ch_pred_params[3]
)
ir_tl <- predict_length(
  age, ir_pred_params[1], ir_pred_params[2], ir_pred_params[3]
)
cbind(age, ch_tl, ir_tl)
```

### Compare the length at age between the two populations. 

Clearly Indian River is growing faster and larger than the Charlotte Harbor. It's also pretty unbiologically sound that at age 0 they are 200mm already given they should be more like 6mm.

## 6. Set up an AIC table:

```{r}
map_columns <- function(v, cols) {
    c <- rep(0, 8)
    i <- 0
    for (col in cols) {
        i <- i + 1
        if (endsWith(col, 'Linf')) {
            j <- 1
        } else if (endsWith(col, 'vbk')) {
            j <- 3
        } else if (endsWith(col, 'tknot')) {
            j <- 5
        } else {
            j <- 7
        }

        if (startsWith(col, 'ch_')) {
            c[j] <- v[i]
        } else if (startsWith(col, 'ir_')) {
            c[j+1] <- v[i]
        } else {
            c[j] <- v[i]
            c[j+1] <- v[i]
        }
    }
    return(c)
}

map_columns(c(1, 3, 2, 3, 4), c('ir_Linf', 'ch_Linf', 'vbk', 'tknot', 'sig'))
```

```{r}
do_likelihood_fit <- function(v, cols, runs) {
  objective <- function(v) {
    c <- map_columns(v, cols)

    ch_Linf <- exp(c[1])
    ch_vbk <- c[3]
    ch_tknot <- c[5]
    ch_sig <- exp(c[7])

    ir_Linf <- exp(c[2])
    ir_vbk <- c[4]
    ir_tknot <- c[6]
    ir_sig <- exp(c[8])

    ch_NLL <- get_likelihood(ch_data$yearsold, ch_data$tl, ch_Linf, ch_vbk, ch_tknot, ch_sig)
    ir_NLL <- get_likelihood(ir_data$yearsold, ir_data$tl, ir_Linf, ir_vbk, ir_tknot, ir_sig)

    NLL <- ch_NLL + ir_NLL
    return(NLL)
  }

  for (i in 1:runs) {
    fit <- optim(v, objective, hessian=T)
    v <- fit$par
  }
  return(fit)
}

summarize_aic <- function(fit, cols) {
  dof <- length(cols)
  nll <- fit$value
  aic <- 2*nll + 2*dof
  row <- c(paste(cols, collapse=','), dof, nll, aic)
  return(row)
}
```

### Determine the log likelihood, the number of parameters, and the AIC for:

#### A model with different parameters ($L_{\infty}, t_0, k,$ and $\sigma$) for each population.

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2,  4.1
) 
cols <- c(
  'ch_Linf', 'ch_vbk', 'ch_tknot', 'ch_sig',
  'ir_Linf', 'ir_vbk', 'ir_tknot', 'ir_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_free <- summarize_aic(fit, cols))
```

#### A model with shared parameters for both populations.  

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7
) 
cols <- c(
  'Linf', 'vbk', 'tknot', 'sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(all_shared <- summarize_aic(fit, cols))
```

### Which hypothesis had the most support based on AIC? 

As in the last lab, all parameters shared has the most support.

## 7. Continue adding to your AIC table.

### Determine the log likelihood function, the log likelihood, the number of parameters, and the AIC for

#### Models that share only one of the VBGM parameters ($L_{\infty}, t_0, k,$ and $\sigma$) for each population (hint: this should result in four additional models for each population).

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  0.2, -1.2,  4.1
) 
cols <- c(
  'Linf', 'ch_vbk', 'ch_tknot', 'ch_sig',
  'ir_vbk', 'ir_tknot', 'ir_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(Linf_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  -1.2,  4.1
) 
cols <- c(
  'ch_Linf', 'vbk', 'ch_tknot', 'ch_sig',
  'ir_Linf', 'ir_tknot', 'ir_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(vbk_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, 4.1
) 
cols <- c(
  'ch_Linf', 'ch_vbk', 'tknot', 'ch_sig',
  'ir_Linf', 'ir_vbk', 'ir_sig'
)
(fit <- do_likelihood_fit(v, cols, 25))
(tknot_shared <- summarize_aic(fit, cols))
```

```{r}
v <- c(
  6.7, 0.2, -1.5,  3.7,
  6.9,  0.2, -1.2
) 
cols <- c(
  'ch_Linf', 'ch_vbk', 'ch_tknot', 'sig',
  'ir_Linf', 'ir_vbk', 'ir_tknot'
)
(fit <- do_likelihood_fit(v, cols, 25))
(sig_shared <- summarize_aic(fit, cols))
```

## 8. Calculate the $\Delta$AIC for all the models you trialed (report your AIC table to answer questions 6, 7, and 8). 

```{r}
aic_table <- data.frame(rbind(
    all_free,
    all_shared,
    Linf_shared,
    vbk_shared,
    tknot_shared,
    sig_shared
))
colnames(aic_table) <- c('cols', 'dof', 'nll', 'aic')

aic_table$aic <- as.numeric(aic_table$aic)
aic_table$delta <- aic_table$aic - min(aic_table$aic)
(aic_table <- aic_table[order(aic_table$delta),])
```

### Which hypothesis had the most support based on AIC? 

Sharing $k$ had the best support.

### What are the top models based on the AIC? 

Sharing $k$, sharing $t_0$, sharing $L_{\infty}$, and all free were the best.

### Do these top models make sense given the biology of Spotted Seatrout? 

Sharing $t_0$ makes sense but the $t_0$ here is certainly unbiological. Growth rates would definitely change from place to place, but I think we're still able to fit reasonably well because $t_0$ and $L_{\infty}$ are still free in the shared $k$ case and we know these are all extremely correlated with one another. Same argument for $L_{\infty}$, I'd expect it to be different but the model is likely taking advantage of the other parameters to still fit. In general across these models the wild $t_0$ makes these rather unbiological.

## 9. Adjust the slot limits for the following scenarios and answer the following questions:

| MinLL | MaxLL | Bias | Over | Under | Parameters |
| ---   | ---   | ---  | ---  | ---   | ---        |
| 300 (0.05) | $\infty$ | $<2$ years old undersampled | $<4$ | N/A | $L_{\infty}=914, k=0.178, t_0=-1.09, \sigma=110.0$ | 
| 0 | 700 (0.05) | $>3$ years old undersampled | $0<age<3$ | $>3$ | $L_{\infty}=667, k=0.335, t_0=-0.39, \sigma=66.2$ |
| 380 (0) | 480 (0) | Almost no sampling for age $>6$ or $<2$, extremely poor sampling across the board. | $<3$ | $>3$ | $L_{\infty}=446, k=0.939, t_0=-1.15, \sigma=26.4$ |
| 380 (0) | 530 (0.03) | No sampling for age $<2$ biased toward middling or smaller fish throughout. | $<3$ | $>3$ | $L_{\infty}=532, k=0.222, t_0=-5.56, \sigma=40.2$ | 

### Across scenarios, which had the biggest impact on $L_{\infty}$? on $k$? on $t_0$? 

- $L_{\infty}$ - Definitely case 3, the total lack of sampling of large fish threw things off entirely.
- $k$ - Likewise case 3.
- $t_0$ - case 4, the complete lack of samples on the low end with samples on the high end gave us a weird almost straight relationship between age and growth.

